---
title: "Measuring Variation"
---

## How Spread Out Is Your Data? {#sec-variation-uncertainty}

Chapter 3 told you where the middle is. This chapter tells you how spread out your data is around that middle.

Two providers might both have a mean incident rate of 10 per month, but one consistently records 9-11 incidents (low variation), while the other swings from 0-20 (high variation). This distinction matters for regulation.

### Why measuring variation matters

- **Consistency vs volatility**: Low variation suggests stable processes, high variation suggests inconsistent performance
- **Comparison context**: A provider 5 incidents above the mean is more concerning if typical variation is ±2 than if it's ±10
- **Method selection**: Different measures suit different data types (quantitative vs qualitative)
- **Outlier detection**: You need to know typical spread to identify unusual values

**Building on Chapter 3**: You know where your data sits (location). This chapter quantifies how variable it is (spread).

## Variation in Quantitative Data

Quantitative data (continuous and discrete) requires measures that use arithmetic. The most common are standard deviation, variance, and interquartile range.

### Standard Deviation and Variance

**Standard deviation (SD)** and **variance** are the same measure, related by a monotonic transformation: variance = SD². They quantify average distance from the mean.

**Key distinction**: Units of measurement and interpretability.

- **Variance**: Squared units (e.g., hours², incidents²). Mathematically convenient but hard to interpret.
- **Standard deviation**: Original units (e.g., hours, incidents). Easy to interpret: "typical distance from mean."

#### When to report which
- **SD for interpretation**: "The standard deviation is 2.5 hours, so most practices are within 2.5 hours of the mean."
- **Variance for calculations**: Many statistical methods (t-tests, ANOVA, regression) work with variance internally.
- **Both are the same information**: If you have one, you have the other. Don't treat them as different measures.

#### Interpretation
- Low SD/variance: Data clustered tightly around mean, consistent performance
- High SD/variance: Data spread out, variable performance
- Example: Mean waiting time = 3.2 hours, SD = 0.8 hours → most practices are between 2.4 and 4.0 hours

#### Why SD/variance matters
- **Z-scoring** (@sec-z-scoring): Quantifies how many SDs a provider is from the mean
- **Control limits in SPC** (@sec-spc-basics): Typically set at ±3 SD from mean
- **Comparing providers**: A provider 1 SD above mean in a low-SD indicator is more unusual than 1 SD above mean in a high-SD indicator

**CQC context**: SD tells you how much natural variation exists across providers. If the SD of incident rates is large, providers naturally vary a lot. If it's small, any deviation is more notable.

### Interquartile Range: Robust to Outliers

**Interquartile range (IQR)** is the distance between the 25th and 75th percentiles (Q3 - Q1). It describes the spread of the middle 50% of your data.

#### Why use IQR instead of SD
- **Robust to outliers**: Extreme values don't affect it
- **Works with skewed data**: Doesn't assume symmetry
- **Easy to communicate**: "The middle half of providers span this range"

#### When to use
- Skewed data where SD is inflated by outliers
- Reporting spread to non-technical audiences
- Box plots (which display IQR visually)

**Example**: IQR = 1.5 hours means the middle 50% of practices have waiting times within a 1.5-hour range. Whether the extreme tail extends to 10 hours or 100 hours doesn't affect this.

**SD vs IQR**: 
- SD uses all data, more efficient with symmetric data
- IQR ignores tails, more robust with skewed data or outliers
- Report both when data is skewed

### Range and Percentile-Based Measures

**Range**: Maximum - Minimum. Simple but heavily influenced by outliers. Rarely useful for CQC work.

**Percentile ranges**: Distance between any two percentiles (e.g., 10th to 90th). More robust than full range.

- **80% range** (P10 to P90): Excludes extreme 10% at each tail
- **90% range** (P5 to P95): Excludes extreme 5% at each tail

#### When to use Describing spread while excluding extreme outliers, especially for skewed distributions.

### Coefficient of Variation: Comparing Across Scales

**Coefficient of Variation (CV)** normalizes SD to make variability comparable across different scales or units.

**Definition**: CV = SD / Mean (often expressed as percentage)

#### Why use CV
- **Compare different indicators**: "Which is more variable, waiting times or incident rates?"
- **Compare different scales**: Providers with very different baseline levels
- **Assess scaling**: Does variability increase proportionally with the mean?

#### Example
- Waiting times: Mean = 3 hours, SD = 0.6 hours → CV = 20%
- Incident rates: Mean = 15 per month, SD = 3 per month → CV = 20%

Both have the same relative variability (20%), even though raw SDs differ.

#### Interpretation
- Low CV (< 15%): Low relative variability, consistent performance
- Moderate CV (15-30%): Moderate relative variability
- High CV (> 30%): High relative variability, inconsistent performance

#### Limitations
- Undefined when mean = 0
- Unstable when mean is close to 0
- Requires ratio scale (meaningful zero point)
- Only works for quantitative data

### Summary: Quantitative Variation Measures

| Measure | Calculation | When to Use | Strengths | Limitations |
|---------|-------------|-------------|-----------|-------------|
| **SD/Variance** | Average distance from mean (variance = SD²) | Symmetric data, statistical methods | Uses all data, efficient | Sensitive to outliers, inflated by skewness |
| **IQR** | Q3 - Q1 | Skewed data, outliers present | Robust to extremes | Ignores tails, less efficient |
| **Range** | Max - Min | Quick check only | Simple | Heavily influenced by extremes |
| **Percentile ranges** | P90 - P10, etc. | Skewed data, excluding extremes | Customizable robustness | Arbitrary percentile choice |
| **CV** | SD / Mean (%) | Comparing across scales | Normalized, comparable | Unstable near zero, ratio scale only |

## Variation in Ordinal Data

Ordinal data (ordered categories) can use quantile-based measures but not arithmetic-based measures.

### What Works for Ordinal Data

**IQR and percentile ranges**: These only require ordering, not arithmetic.

- Example: CQC ratings (Inadequate < Requires Improvement < Good < Outstanding)
- Can calculate: Median rating, IQR of ratings, percentage in each category
- Cannot calculate: Mean rating, SD of ratings (no meaningful arithmetic)

**Percentage in each category**: Simple and interpretable.

- Example: "60% of providers are Good, 30% Outstanding, 10% Requires Improvement"
- Shows distribution without assuming equal spacing between categories

### What Doesn't Work for Ordinal Data

**SD and variance**: Require arithmetic operations. You cannot calculate a meaningful "average distance" from a mean rating when the spacing between categories is undefined.

**Don't do this**: Assign numbers (1=Inadequate, 2=Requires Improvement, 3=Good, 4=Outstanding) and calculate SD. The numbers are arbitrary—why not 1, 2, 4, 10? Different numbering gives different SDs, making it meaningless.

## Variation in Nominal Data

Nominal data (unordered categories) cannot use quantiles or arithmetic. You need heterogeneity measures.

### Heterogeneity Measures for Categorical Data

**Proportion in largest category**: Simple measure of concentration.

- If 90% of providers are in one category, low heterogeneity (homogeneous)
- If categories are evenly split, high heterogeneity (diverse)

**Shannon's Entropy (H)**: Measures uncertainty or diversity in categorical data.

$$H = -\sum_{i=1}^{k} p_i \log_2(p_i)$$

For each category, multiply its proportion by the log₂ of its proportion; sum these products; negate the result. A larger H means more diversity.

Where *p*ᵢ is the proportion in category *i*, and *k* is the number of categories.

#### Interpretation
- H = 0: All observations in one category (no diversity)
- H = log₂(k): Perfectly uniform distribution (maximum diversity)
- Higher H = more heterogeneous/diverse

**Example**: Provider types across a region
- 100% hospitals: H = 0 (no diversity)
- 50% hospitals, 50% care homes: H = 1.0
- 25% each of 4 types: H = 2.0 (maximum for 4 categories)

#### CQC use cases
- Measuring diversity of provider types in a region
- Quantifying heterogeneity of incident categories
- Comparing diversity across regions or time periods

**Simpson's Diversity Index (D)**: Probability that two randomly selected observations are from different categories.

$$D = 1 - \sum_{i=1}^{k} p_i^2$$

#### Interpretation
- D = 0: All observations in one category (no diversity)
- D = (k-1)/k: Perfectly uniform distribution (maximum diversity)
- Higher D = more diverse

**Example**: Same provider type data
- 100% hospitals: D = 0
- 50% hospitals, 50% care homes: D = 0.5
- 25% each of 4 types: D = 0.75

#### When to use which
- **Shannon's Entropy**: More sensitive to rare categories, commonly used in ecology and information theory
- **Simpson's Index**: More intuitive interpretation (probability of difference), less sensitive to rare categories
- **Both are valid**: Choose based on audience and context

::: {.callout-note icon=false}
## Rethinking: Why Heterogeneity Matters for Regulation

Heterogeneity in categorical variables affects how you interpret performance data.

**Example 1: Incident categories**
- Low heterogeneity (90% medication errors): Suggests a specific, targetable problem
- High heterogeneity (evenly split across 10 categories): Suggests systemic quality issues, harder to target

**Example 2: Provider types in a region**
- Low heterogeneity (all care homes): Simpler to benchmark and compare
- High heterogeneity (hospitals, care homes, GP practices, hospices): Need stratified analysis, harder to create fair comparisons

**Example 3: CQC ratings over time**
- Increasing heterogeneity: System becoming more variable, some improving while others decline
- Decreasing heterogeneity: System converging, either improving together or declining together

Measuring heterogeneity helps you understand whether variation is concentrated in specific areas (targetable) or diffuse across many dimensions (systemic).
:::

::: {.callout-warning icon=false}
## Overthinking: Other Normalized Measures

Beyond CV, other normalized measures exist for specific contexts.

### Quartile Coefficient of Dispersion (QCD)

**Definition**: QCD = (Q3 - Q1) / (Q3 + Q1)

**Purpose**: Normalized version of IQR, comparable across scales.

#### Interpretation
- QCD = 0: No spread (all values equal)
- QCD = 1: Maximum theoretical spread
- Higher QCD = more relative variability

#### When to use
- Comparing IQR-based spread across different scales
- When data is skewed (CV not appropriate)
- When you want a robust, normalized dispersion measure

### Normalized Entropy

**Definition**: H_norm = H / log₂(k), where k = number of categories

**Purpose**: Makes Shannon's Entropy comparable across variables with different numbers of categories.

#### Interpretation
- H_norm = 0: All observations in one category
- H_norm = 1: Perfectly uniform distribution
- Scale: Always between 0 and 1, regardless of number of categories

#### When to use
- Comparing heterogeneity across categorical variables with different numbers of categories
- Example: Comparing diversity of incident types (10 categories) vs provider types (4 categories)

### When Normalization Matters

**Comparing across indicators**: "Is waiting time more variable than incident rates?" requires normalized measures (CV, QCD).

**Comparing across time**: If the mean changes over time, raw SD changes too. CV shows whether relative variability changed.

**Comparing across regions**: If regions have different baseline levels, normalized measures show whether relative variability differs.

**When to escalate**: If you're unsure which normalization is appropriate for your comparison, or if stakeholders are confused by normalized measures, seek specialist support.
:::

## Choosing the Right Measure

Match your variation measure to your data type and analytical question.

### By Data Type

| Data Type | Appropriate Measures | Inappropriate Measures |
|-----------|---------------------|------------------------|
| **Continuous** | SD, variance, IQR, range | Entropy, Simpson's Index |
| **Discrete** | SD, variance, IQR, range | Entropy, Simpson's Index |
| **Ordinal** | IQR, percentile ranges, % in each category | SD, variance (no meaningful arithmetic) |
| **Nominal** | Entropy, Simpson's Index, % in largest category | SD, variance, IQR (no ordering) |

### By Data Characteristics

| Characteristic | Preferred Measure | Why |
|----------------|-------------------|-----|
| **Symmetric, no outliers** | SD, variance, CV | Most efficient, uses all data |
| **Outliers present** | IQR, percentile ranges, QCD | Robust to extremes |
| **Comparing scales** | CV, QCD | Normalized measures needed |
| **Small sample** | IQR | Less sensitive to sampling variability |

### By Purpose

| Purpose | Recommended Measure | Example |
|---------|---------------------|---------|
| **Statistical methods** | SD, variance | Input for z-scoring, t-tests, SPC |
| **Robust description** | IQR | Reporting to stakeholders, skewed data |
| **Comparing scales** | CV, QCD | "Which indicator is more variable?" |
| **Categorical diversity** | Entropy, Simpson's Index | "How diverse are incident types?" |
| **Simple communication** | Range, IQR | Non-technical audiences |

## Visualizing Variation

Summary statistics are essential, but visualization reveals patterns numbers alone might miss.

**Histogram**: Shows distribution shape and outliers.

- Good for: Understanding distribution shape before choosing methods
- Reveals: Multimodality, outliers, gaps in data

**Box plot**: Shows median, quartiles (IQR), and outliers in a compact format.

- Good for: Comparing multiple groups side-by-side
- Reveals: Differences in central tendency, spread, and outliers across groups

**Violin plot**: Combines box plot with density plot, showing full distribution shape.

- Good for: Seeing distribution shape while comparing groups
- Reveals: Multimodality and density patterns

**Scatter plot**: Shows relationship between two variables.

- Good for: Exploring whether variables are related
- Reveals: Linear vs nonlinear relationships, clustering, outliers

The example below illustrates why visualisation matters. All four provider types have been assigned similar mean waiting times in their description, but the box plots immediately reveal that **spread differs substantially**. Care homes are tightly clustered; GP practices and mental health trusts show wide variation with potential outliers.

```{r}
#| label: fig-boxplot-variation
#| fig-cap: "Box plots comparing synthetic waiting times (hours) across four CQC-regulated provider types (n = 100 per group, values are illustrative). Care homes show the tightest spread; GP practices and mental health trusts show substantially higher variation, with outliers visible in the mental health trust group."
#| fig-alt: "Horizontal box plots for four provider types. Care Home: median around 3 hours, IQR approximately 2.7–3.3 hours — tightly clustered. Hospital Trust: median around 6 hours, IQR approximately 5.2–6.8 hours — moderate spread. GP Practice: median around 4 hours, IQR approximately 2.3–5.7 hours — wide spread. Mental Health Trust: median around 8 hours, IQR approximately 6–10 hours — widest spread with several high outliers."
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

library(ggplot2)

set.seed(42)
n <- 100
var_df <- data.frame(
  group = factor(
    rep(c("Care Home", "Hospital Trust", "GP Practice", "Mental Health Trust"), each = n),
    levels = c("Care Home", "Hospital Trust", "GP Practice", "Mental Health Trust")
  ),
  waiting_time = c(
    rnorm(n, 3, 0.4),
    rnorm(n, 6, 1.2),
    rnorm(n, 4, 2.5),
    rnorm(n, 8, 3.0)
  )
)
var_df$waiting_time <- pmax(var_df$waiting_time, 0.1)

cqc_palette <- c(
  "Care Home"            = "#006560",
  "Hospital Trust"       = "#007EF5",
  "GP Practice"          = "#A57703",
  "Mental Health Trust"  = "#A83DA4"
)

ggplot(var_df, aes(x = group, y = waiting_time, fill = group)) +
  geom_boxplot(alpha = 0.85, outlier.size = 1.2, outlier.alpha = 0.6) +
  scale_fill_manual(values = cqc_palette, guide = "none") +
  scale_x_discrete(limits = rev(levels(var_df$group))) +
  labs(
    x = NULL,
    y = "Waiting time (hours)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

```{r}
#| label: fig-violin-variation
#| fig-cap: "Violin plots (with embedded median boxes) for the same synthetic data. The violin width shows where values are concentrated. Care homes have a narrow, symmetric distribution; GP practices show a wide, roughly uniform spread; mental health trusts show a right-skewed distribution with a long upper tail."
#| fig-alt: "Violin plots for four provider types. Care Home: a narrow, nearly symmetric violin centred on 3 hours. Hospital Trust: a moderately wide violin around 6 hours, roughly symmetric. GP Practice: a wide, relatively flat violin spanning 0–9 hours around a median of 4 hours. Mental Health Trust: an asymmetric violin extending from around 2 to over 15 hours, with the bulk concentrated between 6 and 10 hours and a long upper tail."
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

ggplot(var_df, aes(x = group, y = waiting_time, fill = group)) +
  geom_violin(alpha = 0.75, trim = TRUE) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.9, outlier.shape = NA) +
  scale_fill_manual(values = cqc_palette, guide = "none") +
  scale_x_discrete(limits = rev(levels(var_df$group))) +
  labs(
    x = NULL,
    y = "Waiting time (hours)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

Always visualize your data before applying statistical methods. Patterns obvious in plots might not be obvious in summary statistics.

## Key Takeaways

::: {.callout-important icon=false}
## Essential Points

1. **Variation is not one thing**: Different data types require different measures. Quantitative → SD/IQR/CV. Ordinal → IQR only. Nominal → Entropy/Simpson's Index.

2. **SD and variance are the same measure**: Related by variance = SD². Use SD for interpretation (original units), variance for calculations (mathematical convenience). Don't treat them as different.

3. **Match measure to data characteristics**: SD for symmetric data without outliers, IQR when outliers present or sample is small.

4. **CV normalizes for scale**: Use CV when comparing variability across different indicators or scales. Makes relative variability comparable.

5. **Ordinal ≠ quantitative**: You can calculate IQR for ordinal data (only needs ordering), but not SD (requires arithmetic). Don't assign arbitrary numbers to categories and calculate SD.

6. **Categorical data needs heterogeneity measures**: Shannon's Entropy and Simpson's Index quantify diversity in nominal data. Higher values = more heterogeneous.

7. **Choose absolute or relative**: Raw measures (SD, IQR) show absolute spread. Normalized measures (CV, QCD) show relative spread. Use relative when comparing across scales.

8. **Visualize before you analyze**: Plots reveal distribution shape, outliers, and patterns that summary statistics might miss.
:::

## Related Approaches

Understanding variation is foundational to every methods chapter. Where to go next:

### Apply variation concepts to methods
- **Monitoring variation over time** → Statistical Process Control (@sec-spc-basics): Control limits are set using SD; distinguishing common cause from special cause variation is built on everything in this chapter.
- **Comparing providers against their peers** → Z-scoring (@sec-z-scoring): Z-scores express a provider's deviation in units of SD — the same SD described here.
- **Investigating unusual values** → @sec-unusual-observations: Defines "unusual" relative to the distribution's spread.
- **Comparing two groups** → t-tests (@sec-t-tests): t-statistics use SD and IQR concepts directly.

### For deeper understanding
- **Shape of the distribution** → @sec-distribution-shape: How skewness and kurtosis affect which variation measure to use.
- **Where the middle sits** → @sec-describing-data: Location and spread together describe a distribution fully.
- **Data quality before analysis** → @sec-qa-principles
