---
title: "Outlier Detection and Treatment"
---

```{r}
#| include: false
library(reticulate)
```

After reading this chapter, you will know how to apply three complementary outlier detection methods (Z-score, Modified Z-score, and IQR), combine their results to prioritise investigation, interpret flagged values in context, and document your decisions in a way that is defensible for QA.

::: {.callout-tip icon=false}
## Problem This Method Solves

You need to systematically identify and investigate data points that stand out from the rest of your dataset. For example:

- A care home reporting staff-to-resident ratios that seem implausibly high or low
- Hospital waiting times that are extreme compared to similar providers
- Incident rates that warrant immediate investigation
- Data quality issues that need identification before formal analysis

Outlier detection provides computational tools to flag unusual observations for investigation, going beyond visual inspection to provide systematic, reproducible criteria.
:::

## What Outlier Detection Does {#sec-outlier-methods}

Outlier detection methods apply statistical rules to identify observations that are unusually far from the bulk of the data. They answer: "Which values are extreme enough to warrant investigation?"

### Key advantages
- Provides systematic, reproducible criteria for flagging unusual values
- Scales to large datasets where visual inspection is impractical
- Offers multiple methods for different data characteristics
- Helps prioritize investigation efforts

### What it does NOT do
- Tell you whether an outlier is an error or genuine extreme value (requires investigation)
- Provide a single "correct" answer (different methods flag different points)
- Replace the conceptual understanding needed before analysis (see @sec-unusual-observations)
- Automatically fix data problems (treatment requires judgment)

## Before You Start

**QA Framework alignment**: Outlier detection most directly supports QA question 3 (Analysis — has the data been appropriately checked before main analysis?) and question 5 (Fitness for Purpose — are the results defensible given the data quality?). Document which detection methods were applied, what investigation was conducted for each flagged observation, and how treatment decisions were made as part of the question 3 record.

::: {.callout-note icon=false}
## Note on Code Examples
The worked examples in this chapter use pseudo-randomly generated data to illustrate the methods. To ensure consistent results between R and Python, the Python code uses the same dataset generated by R (via `r.data_name`). The commented-out Python code shows how you would generate equivalent data independently if needed.
:::

Check these requirements before applying outlier detection:

### Prerequisites
- [ ] **Conceptual foundation**: Read @sec-unusual-observations first to understand what "unusual" means
- [ ] **Data type**: Continuous or count data (categorical data needs different approaches)
- [ ] **Context understanding**: Know what values are plausible for your indicator
- [ ] **Quality checks**: Completed basic data quality checks (see @sec-qa-principles)

### When Outlier Detection May Not Be Appropriate
- **Bimodal distributions**: Two distinct groups may not have "outliers" but different populations
- **Very small samples** (<20): Most methods unreliable
- **Expected extreme values**: If you know some values should be extreme (e.g., specialist units)
- **Time-series data**: Use SPC methods instead (@sec-spc-basics)

### What You Need
- Provider-level or observation-level data
- Understanding of the indicator and what constitutes a plausible range
- Statistical software (R or Python)
- Documentation of why you're detecting outliers (investigation? data quality?)

::: {.callout-note icon=false}
## 📖 Complete Worked Example

**Example 10: Care Home Staffing Investigation** demonstrates outlier detection in practice:

- Identifying extreme staffing ratios using multiple methods
- Investigating whether outliers are errors or genuine extremes
- Handling small provider variation
- Documenting decisions for regulatory defensibility
- Communicating findings to inspectors

**Time**: 2-3 hours | **Complexity**: Medium

[View Example 10: Care Home Staffing Investigation →](../examples/example-10-outliers.qmd)
:::

## Step-by-Step Guide

### Step 1: Visualize Your Data First

Always start with visualization to understand the distribution and identify potential outliers visually.

#### Create distribution plots
- Histogram with density curve to see overall shape
- Box plot to identify outliers using IQR method automatically

#### What to look for
- Is the distribution symmetric or skewed?
- How many potential outliers are visible?
- Are outliers in one tail or both?
- Are outliers isolated points or small clusters?

### Step 2: Apply Multiple Detection Methods

Use 2-3 detection methods and look for consensus. Agreement across methods strengthens the case for investigation.

**Method 1: Z-score**  
Flags observations more than k standard deviations from the mean (typically k = 2.5 or 3).

- **Formula**: $Z = \frac{x - \mu}{\sigma}$
- **When to use**: Data approximately normal, large sample (n > 30)
- **Limitation**: Sensitive to extreme values, assumes normality

**Method 2: Modified Z-score**  
Uses median and median absolute deviation (MAD) instead of mean/SD. More robust to outliers.

- **Formula**: $Z_{mod} = \frac{0.6745(x - median)}{MAD}$
- **Threshold**: Typically |Z_mod| > 3.5
- **When to use**: Data skewed or contains outliers that distort mean/SD
- **Advantage**: Robust to outliers, works with non-normal data

**Method 3: IQR Method (Tukey's Fences)**  
Uses interquartile range to define fences. The method box plots use.

- **Fences**: $Q1 - 1.5 \times IQR$ and $Q3 + 1.5 \times IQR$
- **When to use**: Non-normal data, want method matching box plots
- **Note**: Can flag many points in skewed distributions

### Step 3: Compare Method Results

Count how many methods flag each observation:
- **3 methods agree**: Highest priority for investigation
- **2 methods agree**: High priority
- **1 method only**: May be method-specific artifact

### Step 4: Investigate Flagged Observations

Detection is just the first step. Every flagged observation requires investigation.

#### For each flagged observation, ask

1. **Data quality**: Could this be an error?
   - Implausible value (negative, impossible)?
   - Wrong unit or scale?
   - Missing data coded as extreme value?

2. **Context**: Why might this be genuinely extreme?
   - Small provider (high variation expected)?
   - Specialist service (different population)?
   - Known event (incident, merger, closure)?

3. **Consistency**: Do related values support this?
   - Other time periods for same provider?
   - Related indicators?
   - Inspection reports or narrative data?

### Step 5: Decide on Treatment

Based on investigation, choose appropriate action:

**Option 1: Keep as genuine extreme** → Verified genuine, provides important information  
**Option 2: Remove from analysis** → Confirmed data error, provider not comparable  
**Option 3: Adjust or transform** → Systematic bias identified  
**Option 4: Separate analysis** → Distinct subgroups identified

Document all decisions in a decision log for QA purposes.

## Worked Example: Care Home Staffing Ratios

### Scenario

You're analysing staffing ratios (qualified staff per 10 residents) across 80 care homes. You need to identify extreme values that warrant investigation before conducting your main analysis.

**Data**: Simulated staffing ratios (most homes 2.5-4.5 staff per 10 residents, with some genuine extremes and data errors)

### Step 1: Generate and Visualize Data

::: {.panel-tabset group="language"}
## R
```{r}
#| label: outlier-data
#| echo: true
set.seed(123)

# Generate 80 care homes with staffing ratios
care_homes <- data.frame(
  home_id = sprintf("CH%03d", 1:80),
  staff_per_10 = c(
    rnorm(65, mean = 3.2, sd = 0.7),      # Normal homes
    rnorm(5, mean = 5.5, sd = 0.3),        # High staffing (specialist units)
    rnorm(3, mean = 1.5, sd = 0.2),        # Low staffing
    c(0.02, 0.05, 9.8, 11.2, 0, 8.5, 7.2)  # Data errors and extremes
  )
)

# Ensure non-negative
care_homes$staff_per_10 <- pmax(care_homes$staff_per_10, 0)

# Display summary
summary(care_homes$staff_per_10)
```

```{r}
#| label: outlier-viz
#| echo: true
#| fig-width: 10
#| fig-height: 4

# Visualize distribution
par(mfrow = c(1, 2))

# Histogram
hist(care_homes$staff_per_10, breaks = 20, 
     main = "Distribution of Staffing Ratios",
     xlab = "Staff per 10 Residents",
     col = "lightblue", border = "white")

# Box plot
boxplot(care_homes$staff_per_10, 
        main = "Box Plot Showing Outliers",
        ylab = "Staff per 10 Residents",
        col = "lightblue", outcol = "red", outpch = 19)
abline(h = c(2.5, 4.0), lty = 2, col = "darkgreen")
text(1.3, 2.5, "Min guideline", col = "darkgreen", cex = 0.8)
text(1.3, 4.0, "Good practice", col = "darkgreen", cex = 0.8)
```

## Python
```{python}
#| label: outlier-data-py
#| echo: true

# Use R's data for consistency
care_homes = r.care_homes.copy()

# Display summary
print(care_homes['staff_per_10'].describe())
```

```{python}
#| label: outlier-viz-py
#| echo: true
#| fig-width: 10
#| fig-height: 4

import matplotlib.pyplot as plt

fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(10, 4))

# Histogram
ax1.hist(care_homes['staff_per_10'], bins=20, color='lightblue', edgecolor='white')
ax1.set_xlabel('Staff per 10 Residents')
ax1.set_ylabel('Frequency')
ax1.set_title('Distribution of Staffing Ratios')
ax1.grid(alpha=0.3)

# Box plot
ax2.boxplot(care_homes['staff_per_10'], patch_artist=True,
            boxprops=dict(facecolor='lightblue'),
            flierprops=dict(marker='o', markerfacecolor='red', markersize=8))
ax2.axhline(2.5, linestyle='--', color='darkgreen', alpha=0.6)
ax2.axhline(4.0, linestyle='--', color='darkgreen', alpha=0.6)
ax2.set_ylabel('Staff per 10 Residents')
ax2.set_title('Box Plot Showing Outliers')
ax2.set_xticks([])
ax2.grid(alpha=0.3, axis='y')

plt.tight_layout()
plt.show()
```
:::

**Observation**: Right-skewed distribution with several extreme values in both tails. Box plot shows ~10 potential outliers.

### Step 2: Apply Detection Methods

::: {.panel-tabset group="language"}
## R
```{r}
#| label: outlier-detect
#| echo: true

# Z-score method
care_homes$z_score <- scale(care_homes$staff_per_10)[,1]
care_homes$outlier_z <- abs(care_homes$z_score) > 2.5

# Modified Z-score method
median_val <- median(care_homes$staff_per_10)
mad_val <- mad(care_homes$staff_per_10)
care_homes$modified_z <- 0.6745 * (care_homes$staff_per_10 - median_val) / mad_val
care_homes$outlier_modz <- abs(care_homes$modified_z) > 3.5

# IQR method
Q1 <- quantile(care_homes$staff_per_10, 0.25)
Q3 <- quantile(care_homes$staff_per_10, 0.75)
IQR_val <- IQR(care_homes$staff_per_10)
lower_fence <- Q1 - 1.5 * IQR_val
upper_fence <- Q3 + 1.5 * IQR_val
care_homes$outlier_iqr <- care_homes$staff_per_10 < lower_fence | 
                          care_homes$staff_per_10 > upper_fence

# Summary table
detection_summary <- data.frame(
  Method = c("Z-score (|z| > 2.5)", "Modified Z-score (|z| > 3.5)", "IQR (k=1.5)"),
  Outliers = c(sum(care_homes$outlier_z), 
               sum(care_homes$outlier_modz), 
               sum(care_homes$outlier_iqr))
)
print(detection_summary, row.names = FALSE)
cat("\nIQR Fences: [", round(lower_fence, 2), ",", round(upper_fence, 2), "]\n")
```

## Python
```{python}
#| label: outlier-detect-py
#| echo: true

from scipy import stats
from scipy.stats import median_abs_deviation

# Z-score method
care_homes['z_score'] = stats.zscore(care_homes['staff_per_10'])
care_homes['outlier_z'] = abs(care_homes['z_score']) > 2.5

# Modified Z-score method
median_val = care_homes['staff_per_10'].median()
mad_val = median_abs_deviation(care_homes['staff_per_10'])
care_homes['modified_z'] = 0.6745 * (care_homes['staff_per_10'] - median_val) / mad_val
care_homes['outlier_modz'] = abs(care_homes['modified_z']) > 3.5

# IQR method
Q1 = care_homes['staff_per_10'].quantile(0.25)
Q3 = care_homes['staff_per_10'].quantile(0.75)
IQR_val = Q3 - Q1
lower_fence = Q1 - 1.5 * IQR_val
upper_fence = Q3 + 1.5 * IQR_val
care_homes['outlier_iqr'] = ((care_homes['staff_per_10'] < lower_fence) | 
                             (care_homes['staff_per_10'] > upper_fence))

# Summary table
import pandas as pd
detection_summary = pd.DataFrame({
    'Method': ['Z-score (|z| > 2.5)', 'Modified Z-score (|z| > 3.5)', 'IQR (k=1.5)'],
    'Outliers': [care_homes['outlier_z'].sum(), 
                 care_homes['outlier_modz'].sum(), 
                 care_homes['outlier_iqr'].sum()]
})
print(detection_summary.to_string(index=False))
print(f"\nIQR Fences: [{lower_fence:.2f}, {upper_fence:.2f}]")
```
:::

**Analysis**: Modified Z-score is more conservative (fewer false positives) due to robustness. IQR matches box plot visualization.

### Step 3: Compare Methods

::: {.panel-tabset group="language"}
## R
```{r}
#| label: outlier-compare
#| echo: true

# Count flags per home
care_homes$num_flags <- care_homes$outlier_z + care_homes$outlier_modz + 
                        care_homes$outlier_iqr

# Agreement summary
cat("Homes flagged by: ", paste(names(table(care_homes$num_flags)), "method(s) =", 
                                 table(care_homes$num_flags)), "\n")

# High-priority cases (2+ methods)
flagged <- care_homes[care_homes$num_flags >= 2, 
                      c("home_id", "staff_per_10", "z_score", "modified_z", "num_flags")]
flagged <- flagged[order(-abs(flagged$z_score)), ]

cat("\nHigh-priority homes (2+ methods agree):\n")
print(flagged, row.names = FALSE)
```

## Python
```{python}
#| label: outlier-compare-py
#| echo: true

# Count flags per home
care_homes['num_flags'] = (care_homes['outlier_z'].astype(int) + 
                           care_homes['outlier_modz'].astype(int) + 
                           care_homes['outlier_iqr'].astype(int))

# Agreement summary
agreement = care_homes['num_flags'].value_counts().sort_index()
print("Homes flagged by:", ", ".join([f"{k} method(s) = {v}" for k, v in agreement.items()]))

# High-priority cases
flagged = care_homes[care_homes['num_flags'] >= 2][
    ['home_id', 'staff_per_10', 'z_score', 'modified_z', 'num_flags']
].sort_values('z_score', key=abs, ascending=False)

print("\nHigh-priority homes (2+ methods agree):")
print(flagged.to_string(index=False))
```
:::

**Interpretation**: Homes with very low (<1) or very high (>7) ratios flagged by multiple methods are highest priority for investigation. Agreement across methods strengthens case that these warrant attention.

## Common Pitfalls

### Pitfall 1: Mechanical application without investigation

Detecting outliers doesn't tell you what to do with them. Every flagged observation requires investigation to determine whether it's an error, genuine extreme, or contextual issue.

### Pitfall 2: Using wrong method for data type

Z-scores assume normality. For highly skewed data, use modified Z-scores or IQR. Don't force methods onto data that violates assumptions.

### Pitfall 3: Removing outliers to "improve" results

Never remove observations just to make your analysis tidier or improve model fit. Outliers contain information. Only remove when you have evidence of errors.

### Pitfall 4: Not documenting decisions

"We removed 5 outliers" is insufficient for QA. Document which observations, which methods, what investigation revealed, and why you chose your treatment.

### Pitfall 5: Ignoring small numbers

Small providers naturally show higher variation. An extreme value from a 15-bed home is less surprising than from a 100-bed home. Consider denominators when interpreting outliers.

### Pitfall 6: Single method only

Using only one detection method misses the bigger picture. Apply 2-3 methods and look for consensus. Agreement across methods strengthens the case for investigation.

::: {.callout-note icon=false}
## 🤔 Rethinking: Outliers Are Data, Not Problems

The instinct when encountering an outlier is to remove it — to "clean" the data. Resist this. Outliers are often the most important observations in a regulatory dataset.

**Outliers can be**:
- **Data errors**: Wrong units, transcription mistakes, impossible values — these should be corrected or removed with a documented rationale
- **Genuine extremes**: A care home that is genuinely understaffed, or a practice with an exceptionally high referral rate — these are exactly what CQC needs to investigate
- **Contextual differences**: A specialist unit serving a different population — these require separate analysis or adjustment, not removal

**The regulatory risk of over-removing**: If you remove outliers to produce a "cleaner" distribution, you may be removing the very providers that need regulatory attention. A bias towards false negatives (missing problems) is more serious in a regulatory context than a bias towards false positives (over-investigating).

**The right approach**: Flag, investigate, document. Keep outliers unless you can demonstrate they are errors. Report sensitivity analyses showing whether your conclusions change with or without them.
:::

::: {.callout-warning icon=false}
## ⚙️ Overthinking: Beyond Univariate Detection

The methods in this chapter detect outliers on a single variable at a time. But a provider can be unusual in a way that only becomes apparent when you look at multiple variables together.

**Example**: A care home with a staffing ratio of 3.1 (within normal range), an incident rate of 2.8 per resident (within normal range), and a complaint rate of 0.15 per resident (within normal range) — but the combination of all three moderately elevated indicators might signal systematic risk that no single indicator detects.

**Multivariate outlier methods**:
- **Mahalanobis distance**: Measures how far a point is from the multivariate centre, accounting for correlations between variables. A point can be unusual in multivariate space while normal on each individual dimension.
- **Isolation Forest**: Machine learning approach that detects outliers by how quickly they are "isolated" in random splits. Works well with many variables.
- **Local Outlier Factor (LOF)**: Identifies points that are locally unusual relative to their neighbours, rather than globally unusual relative to all data.

**When to use**: When you have 3+ indicators and suspect that unusual combinations matter, not just single extreme values. These methods are computationally straightforward but require care in interpretation.

For multivariate outlier analysis in a regulatory context, raise this with the Quantitative Analytics and Statistics Guild (the Guild) before proceeding — the choice of method and threshold requires specialist judgment.
:::

## Documenting Your Analysis

For QA purposes, record:

### Detection approach
- Which methods applied and why chosen
- Thresholds used and rationale
- Number of outliers flagged by each method
- Agreement/disagreement across methods

### Investigation process
- Which observations investigated
- Evidence gathered (provider contact, related data, narrative)
- Findings for each case
- Time spent on investigation

### Treatment decisions
- Which outliers kept, which removed, which adjusted
- Justification for each decision
- How treatment affects conclusions
- Sensitivity analysis (do conclusions change?)

### Evidence to retain
- Distribution plots (before and after)
- Detection method outputs with thresholds
- Investigation log with documented decisions
- Final dataset with exclusion flags and reasons

See @sec-qa-principles for full QA documentation requirements and templates.

## Related Approaches

### For different data characteristics
- **Multivariate outliers** → See Overthinking box above; raise with the Guild for implementation
- **Time-series outliers** → SPC methods (@sec-spc-basics)
- **Clustering-based detection** → DBSCAN, Isolation Forest — raise with the Guild

### For foundational understanding
- **Conceptual framework** → @sec-unusual-observations (read first!)
- **Small provider variation** → @sec-variation-uncertainty
- **Data quality checks** → @sec-qa-principles

### For related analyses
- **Provider comparison** → Z-scoring (@sec-z-scoring) handles expected variation
- **Missing data** → @sec-missing-data-foundations (outliers vs missing different issues)

## When to Seek Support from the Guild

Raise your analysis with the Quantitative Analytics and Statistics Guild (the Guild) at a Quant Clinic session when:

- **Multivariate outliers needed**: Outliers across multiple variables simultaneously (see Overthinking box)
- **Advanced methods required**: Isolation Forest, Local Outlier Factor, or other machine learning approaches
- **High-stakes decisions**: Outlier treatment could materially affect regulatory action
- **Methodological challenge**: A stakeholder questions your outlier treatment approach
- **Unusual data structure**: Hierarchical data, spatial data, or other complex structures
- **Systematic patterns**: Many outliers suggesting data quality or population issues
- **Resource constraints**: Too many outliers to investigate manually

## Key Takeaways

::: {.callout-important icon=false}
## Essential Points

1. **Detection ≠ decision**: Flagging outliers is just the first step; investigation determines what to do
2. **Multiple methods essential**: Use 2-3 methods and look for consensus across approaches
3. **Context is critical**: An outlier in one context may be expected in another
4. **Document everything**: Which observations flagged, investigated, and treated—with justification
5. **Visual inspection first**: Always plot your data before applying detection methods
6. **Never remove mechanically**: Only remove outliers when you have evidence they're errors
7. **Consider small numbers**: Small providers show more variation—don't penalize natural spread

Outlier detection is about systematically prioritizing investigation, not automatically removing inconvenient data points.
:::

## Further Reading

### Internal CQC Resources

- **QA Framework**: Documentation standards for outlier investigation and treatment
- **Getting help**: See @sec-getting-started for how to access support
- **Data Quality Guidance**: Distinguishing errors from genuine extremes

### External Guidance

- **Tukey JW** (1977). *Exploratory Data Analysis*. Addison-Wesley. (Classic text on outlier detection including box plots and fences)
- **Iglewicz B, Hoaglin DC** (1993). "How to Detect and Handle Outliers." *ASQC Basic References in Quality Control* vol. 16. (Modified Z-score method)
- **Rousseeuw PJ, Hubert M** (2011). "Robust statistics for outlier detection." *Wiley Interdisciplinary Reviews: Data Mining and Knowledge Discovery*, 1(1), 73-79.

### Online Resources

- **NIST Engineering Statistics Handbook**: Section on outlier detection (freely available, comprehensive)
- **R Documentation**: `?boxplot.stats` for Tukey's method implementation
- **Python SciPy**: `scipy.stats.zscore` and `scipy.stats.median_abs_deviation` documentation