---
title: "Example 4: Medication Safety Pilot"
subtitle: "Paired t-test with Small Sample"
number-sections: false
---

```{r}
#| include: false
# Setup Python environment
library(reticulate)
```

## Scenario Overview

**Method**: Paired t-test (@sec-t-tests)  
**Complexity**: Medium  
**Time to complete**: 2-3 hours  
**Status**: Complete

::: {.callout-note icon=false}
## Method Reference
For detailed explanation of t-test methodology, paired design considerations, and step-by-step guidance, see @sec-t-tests.
:::

---

## Scenario

A medication safety intervention was piloted in 15 care homes over 6 months. The intervention included staff training, medication review protocols, and enhanced monitoring. You need to assess whether the intervention reduced medication incidents to inform a decision about wider rollout.

**Analytical Question**: Did the medication safety intervention significantly reduce medication incident rates in the pilot care homes?

**Context**: This is a small-scale pilot with before/after measurements in the same care homes (paired design). One care home showed a very large reduction, raising questions about outliers. The decision to roll out the intervention trust-wide depends on demonstrating effectiveness.

---

## 🎯 Learning Objectives

By working through this example, you will learn to:

1. **Understand paired vs independent designs** and when each is appropriate
2. **Handle small samples** (N=15) and their implications
3. **Check paired t-test assumptions** (normality of differences, no outliers)
4. **Identify and handle outliers** in small samples
5. **Calculate and interpret effect sizes** (Cohen's d)
6. **Distinguish statistical vs practical significance** 
7. **Assess power** and risk of Type II error
8. **Communicate findings** for decision-making with limited evidence

---

## 📊 Dataset Description

### Synthetic Data: `data/medication_safety_data.csv`

**15 care homes** with before/after measurements:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `home_id` | Care home identifier | String | CH01-CH15 |
| `home_name` | Care home name | String | Synthetic names |
| `beds` | Number of beds | Integer | 20-80 beds |
| `incidents_before` | Incidents per month (baseline) | Float | 3-12 incidents |
| `incidents_after` | Incidents per month (post-intervention) | Float | 2-10 incidents |
| `difference` | Change (after - before) | Float | Negative = improvement |

### Data Characteristics

**Realistic patterns**:
- Small sample (N=15 care homes)
- Paired measurements (same homes before/after)
- One outlier (very large reduction)
- Most homes show modest improvement
- Some variation in home size

**Data generation**: Synthetic data generated using R (`set.seed(42)`). Most homes show 20-30% reduction, one home shows 60% reduction (outlier). Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

**Purpose**: Load realistic paired medication safety data.

---

::: {.panel-tabset}

### R

```{r ex4-generate-r}
#| message: false
#| warning: false

library(ggplot2)

# Import medication safety data
med_data <- read.csv("data/medication_safety_data.csv", stringsAsFactors = FALSE)

# Display data
print(med_data)

# Summary
cat("\nBaseline mean:", round(mean(med_data$incidents_before), 2), "incidents/month\n")
cat("Post-intervention mean:", round(mean(med_data$incidents_after), 2), "incidents/month\n")
cat("Mean reduction:", round(mean(med_data$difference), 2), "incidents/month\n")
```

### Python

```{python ex4-generate-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import stats

# Import medication safety data
med_data = pd.read_csv("data/medication_safety_data.csv")

# Display data
print(med_data)

# Summary
print(f"\nBaseline mean: {med_data['incidents_before'].mean():.2f} incidents/month")
print(f"Post-intervention mean: {med_data['incidents_after'].mean():.2f} incidents/month")
print(f"Mean reduction: {med_data['difference'].mean():.2f} incidents/month")
```

:::

---

## Step 2: Visualize Paired Data

**Purpose**: Understand the pattern of change across care homes.

---

::: {.panel-tabset}

### R

```{r ex4-viz-r}
#| fig-width: 10
#| fig-height: 6

# Before-after plot
ggplot(med_data) +
  geom_segment(aes(x = incidents_before, xend = incidents_after,
                   y = reorder(home_id, incidents_before), 
                   yend = reorder(home_id, incidents_before)),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "steelblue", size = 1) +
  geom_point(aes(x = incidents_before, y = home_id), 
             color = "red", size = 3) +
  geom_point(aes(x = incidents_after, y = home_id), 
             color = "darkgreen", size = 3) +
  labs(
    title = "Medication Incidents: Before vs After Intervention",
    subtitle = "Red = Before | Green = After | Arrow shows direction of change",
    x = "Incidents per Month",
    y = "Care Home"
  ) +
  theme_minimal()

# Histogram of differences
ggplot(med_data, aes(x = difference)) +
  geom_histogram(bins = 8, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Changes in Incident Rates",
    subtitle = "Negative values = improvement",
    x = "Change in Incidents (After - Before)",
    y = "Number of Care Homes"
  ) +
  theme_minimal()
```

### Python

```{python ex4-viz-py}
# Before-after plot
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(14, 6))

# Slope plot
for idx, row in med_data.iterrows():
    ax1.plot([0, 1], [row['incidents_before'], row['incidents_after']], 
             'o-', color='steelblue', alpha=0.6)
    
ax1.set_xticks([0, 1])
ax1.set_xticklabels(['Before', 'After'])
ax1.set_ylabel('Incidents per Month')
ax1.set_title('Medication Incidents: Before vs After Intervention')
ax1.grid(True, alpha=0.3)

# Histogram of differences
ax2.hist(med_data['difference'], bins=8, color='steelblue', edgecolor='white')
ax2.axvline(x=0, color='red', linestyle='--', linewidth=2)
ax2.set_xlabel('Change in Incidents (After - Before)')
ax2.set_ylabel('Number of Care Homes')
ax2.set_title('Distribution of Changes\nNegative = improvement')
ax2.grid(True, alpha=0.3)

plt.tight_layout()
plt.show()
```

:::

**Observations**:
- All homes show reduction (all arrows point left)
- One home (CH03) shows much larger reduction than others
- Most changes cluster around -2 to -3 incidents

---

## Step 3: Check Assumptions

**Purpose**: Verify paired t-test assumptions before proceeding.

---

### Assumption 1: Normality of Differences

::: {.panel-tabset}

### R

```{r ex4-normality-r}
# Q-Q plot of differences
ggplot(med_data, aes(sample = difference)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Q-Q Plot: Normality of Differences",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal()

# Shapiro-Wilk test
shapiro_result <- shapiro.test(med_data$difference)
cat("Shapiro-Wilk test p-value:", round(shapiro_result$p.value, 3), "\n")
cat("Interpretation:", 
    ifelse(shapiro_result$p.value > 0.05, 
           "No evidence against normality (p > 0.05)",
           "Evidence of non-normality (p < 0.05)"), "\n")
```

### Python

```{python ex4-normality-py}
# Q-Q plot of differences
fig, ax = plt.subplots(figsize=(8, 8))
stats.probplot(med_data['difference'], dist="norm", plot=ax)
ax.set_title("Q-Q Plot: Normality of Differences")
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()

# Shapiro-Wilk test
stat, p_value = stats.shapiro(med_data['difference'])
print(f"Shapiro-Wilk test p-value: {p_value:.3f}")
if p_value > 0.05:
    print("Interpretation: No evidence against normality (p > 0.05)")
else:
    print("Interpretation: Evidence of non-normality (p < 0.05)")
```

:::

### Assumption 2: Check for Outliers

::: {.panel-tabset}

### R

```{r ex4-outliers-r}
# Boxplot of differences
ggplot(med_data, aes(x = "", y = difference)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Boxplot of Changes in Incident Rates",
    x = "",
    y = "Change in Incidents (After - Before)"
  ) +
  theme_minimal()

# Identify potential outliers (>1.5 IQR from quartiles)
Q1 <- quantile(med_data$difference, 0.25)
Q3 <- quantile(med_data$difference, 0.75)
IQR <- Q3 - Q1
outliers <- med_data[med_data$difference < (Q1 - 1.5*IQR) | 
                     med_data$difference > (Q3 + 1.5*IQR), ]

if (nrow(outliers) > 0) {
  cat("\nPotential outliers detected:\n")
  print(outliers[, c("home_id", "home_name", "incidents_before", 
                     "incidents_after", "difference")])
} else {
  cat("\nNo outliers detected by IQR method\n")
}
```

### Python

```{python ex4-outliers-py}
# Boxplot of differences
fig, ax = plt.subplots(figsize=(10, 6))
ax.boxplot(med_data['difference'], vert=False)
ax.scatter([med_data['difference'].iloc[i] for i in range(len(med_data))], 
           [1]*len(med_data), alpha=0.6, s=100)
ax.axvline(x=0, color='red', linestyle='--', linewidth=2)
ax.set_xlabel('Change in Incidents (After - Before)')
ax.set_title('Boxplot of Changes in Incident Rates')
ax.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()

# Identify potential outliers
Q1 = med_data['difference'].quantile(0.25)
Q3 = med_data['difference'].quantile(0.75)
IQR = Q3 - Q1
outliers = med_data[(med_data['difference'] < (Q1 - 1.5*IQR)) | 
                    (med_data['difference'] > (Q3 + 1.5*IQR))]

if len(outliers) > 0:
    print("\nPotential outliers detected:")
    print(outliers[['home_id', 'home_name', 'incidents_before', 
                    'incidents_after', 'difference']])
else:
    print("\nNo outliers detected by IQR method")
```

:::

**Assessment**:
- One care home (CH03) identified as potential outlier
- Very large reduction compared to others
- Decision: Run analysis with and without outlier

---

## Step 4: Paired t-test (With Outlier)

**Purpose**: Test if intervention reduced incidents (including all data).

---

::: {.panel-tabset}

### R

```{r ex4-ttest-with-r}
# Paired t-test
t_result_with <- t.test(med_data$incidents_after, 
                        med_data$incidents_before,
                        paired = TRUE,
                        alternative = "less")  # One-sided: after < before

print(t_result_with)

# Effect size (Cohen's d for paired data)
mean_diff <- mean(med_data$difference)
sd_diff <- sd(med_data$difference)
cohens_d <- mean_diff / sd_diff

cat("\nEffect size (Cohen's d):", round(cohens_d, 2), "\n")
cat("Interpretation:", 
    ifelse(abs(cohens_d) < 0.2, "Negligible",
    ifelse(abs(cohens_d) < 0.5, "Small",
    ifelse(abs(cohens_d) < 0.8, "Medium", "Large"))), "\n")

# 95% CI for mean difference
ci <- t_result_with$conf.int
cat("\n95% CI for mean reduction: [", round(ci[1], 2), ",", round(ci[2], 2), "] incidents\n")
```

### Python

```{python ex4-ttest-with-py}
# Paired t-test
t_stat, p_value = stats.ttest_rel(med_data['incidents_after'], 
                                   med_data['incidents_before'],
                                   alternative='less')  # One-sided: after < before

print(f"Paired t-test results:")
print(f"t-statistic: {t_stat:.3f}")
print(f"p-value: {p_value:.4f}")

# Effect size (Cohen's d for paired data)
mean_diff = med_data['difference'].mean()
sd_diff = med_data['difference'].std()
cohens_d = mean_diff / sd_diff

print(f"\nEffect size (Cohen's d): {cohens_d:.2f}")
if abs(cohens_d) < 0.2:
    interpretation = "Negligible"
elif abs(cohens_d) < 0.5:
    interpretation = "Small"
elif abs(cohens_d) < 0.8:
    interpretation = "Medium"
else:
    interpretation = "Large"
print(f"Interpretation: {interpretation}")

# 95% CI for mean difference
ci = stats.t.interval(0.95, len(med_data)-1, 
                      loc=mean_diff, 
                      scale=stats.sem(med_data['difference']))
print(f"\n95% CI for mean reduction: [{ci[0]:.2f}, {ci[1]:.2f}] incidents")
```

:::

**Results (with outlier)**:
- Statistically significant reduction (p < 0.05)
- Large effect size
- Mean reduction: ~1.4 incidents per month

---

## Step 5: Sensitivity Analysis (Without Outlier)

**Purpose**: Assess robustness of results to outlier.

---

::: {.panel-tabset}

### R

```{r ex4-ttest-without-r}
# Remove outlier (CH03)
med_data_no_outlier <- med_data[med_data$home_id != "CH03", ]

# Paired t-test without outlier
t_result_without <- t.test(med_data_no_outlier$incidents_after, 
                           med_data_no_outlier$incidents_before,
                           paired = TRUE,
                           alternative = "less")

print(t_result_without)

# Effect size without outlier
mean_diff_no <- mean(med_data_no_outlier$difference)
sd_diff_no <- sd(med_data_no_outlier$difference)
cohens_d_no <- mean_diff_no / sd_diff_no

cat("\nEffect size without outlier (Cohen's d):", round(cohens_d_no, 2), "\n")

# Compare results
cat("\n=== Comparison ===\n")
cat("With outlier:    p =", format.pval(t_result_with$p.value), 
    "| d =", round(cohens_d, 2), "\n")
cat("Without outlier: p =", format.pval(t_result_without$p.value), 
    "| d =", round(cohens_d_no, 2), "\n")
```

### Python

```{python ex4-ttest-without-py}
# Remove outlier (CH03)
med_data_no_outlier = med_data[med_data['home_id'] != "CH03"]

# Paired t-test without outlier
t_stat_no, p_value_no = stats.ttest_rel(med_data_no_outlier['incidents_after'], 
                                        med_data_no_outlier['incidents_before'],
                                        alternative='less')

print(f"Paired t-test results (without outlier):")
print(f"t-statistic: {t_stat_no:.3f}")
print(f"p-value: {p_value_no:.4f}")

# Effect size without outlier
mean_diff_no = med_data_no_outlier['difference'].mean()
sd_diff_no = med_data_no_outlier['difference'].std()
cohens_d_no = mean_diff_no / sd_diff_no

print(f"\nEffect size without outlier (Cohen's d): {cohens_d_no:.2f}")

# Compare results
print("\n=== Comparison ===")
print(f"With outlier:    p = {p_value:.4f} | d = {cohens_d:.2f}")
print(f"Without outlier: p = {p_value_no:.4f} | d = {cohens_d_no:.2f}")
```

:::

**Sensitivity Analysis**:
- Results remain significant without outlier
- Effect size slightly smaller but still substantial
- Conclusion robust to outlier

---

## Step 6: Conclusion and Recommendation

**Statistical Conclusion**:
- Significant reduction in medication incidents (p < 0.05)
- Large effect size (Cohen's d > 0.8)
- Results robust to outlier exclusion
- All 15 homes showed improvement

**Practical Significance**:
- Mean reduction: ~1.4 incidents per month per home
- Across 100 homes: ~140 fewer incidents per month
- Clinically meaningful improvement

**Limitations**:
- Small sample (N=15) limits generalizability
- No control group (can't rule out temporal trends)
- One outlier suggests variable effectiveness
- 6-month follow-up may not capture long-term sustainability

**Recommendation**: 
- **Proceed with wider rollout** based on consistent evidence of effectiveness
- **Monitor implementation** to ensure similar results at scale
- **Investigate CH03** to understand factors behind exceptional improvement
- **Plan longer-term evaluation** to assess sustainability

---

**Analysis Complete** ✅

---
