---
title: "Example 3: Mental Health Wait Times"
subtitle: "Z-scoring with Skewed Data"
number-sections: false
---

```{r}
#| include: false
# Setup Python environment
library(reticulate)
```

## Scenario Overview

**Method**: Z-scoring with transformation (@sec-z-scoring)  
**Complexity**: Medium-High  
**Time to complete**: 2-3 hours  
**Status**: ✅ Complete

::: {.callout-note icon=false}
## 📖 Method Reference
For detailed explanation of z-scoring methodology, handling skewed data, and step-by-step guidance, see @sec-z-scoring.
:::

---

## 📋 Scenario

You are analyzing wait times from assessment to treatment across 150 community mental health services using the 2024 Community Mental Health Survey. Most services have wait times under 3 months, but some have waits exceeding 6 months, creating a highly skewed distribution.

**Analytical Question**: Which mental health services have wait times that are statistically significantly worse than the national average, accounting for the skewed distribution?

**Context**: Mental health wait times are a key quality indicator. The data shows substantial right skew - most services perform well, but a minority have very long waits. Standard z-scoring assumes normality, which may not hold here.

---

## 🎯 Learning Objectives

By working through this example, you will learn to:

1. **Identify non-normal distributions** through visual and statistical checks
2. **Understand when z-scoring assumptions fail** and why it matters
3. **Apply data transformations** (log transformation) to reduce skew
4. **Compare approaches**: raw z-scores vs transformed z-scores
5. **Interpret results** when assumptions are violated
6. **Conduct sensitivity analysis** to assess robustness
7. **Communicate findings** about skewed data to stakeholders
8. **Document analytical decisions** when methods require adaptation

---

## 📊 Dataset Description

### Synthetic Data: `data/mental_health_wait_times.csv`

**150 mental health services** with the following variables:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `service_code` | Unique service identifier | String | Format: MHS001-MHS150 |
| `service_name` | Service name | String | Synthetic names |
| `trust_name` | NHS Trust | String | 50 trusts |
| `region` | NHS England region | String | 7 regions |
| `wait_time_days` | Median wait time (days) | Integer | 30-250 days |
| `survey_responses` | Survey respondents | Integer | 50-300 |
| `deprivation_score` | Area deprivation (IMD) | Float | 10-40 |

### Data Characteristics

**Realistic issues included**:
- Strong positive skew (right tail)
- Most services 60-90 days, some >180 days
- No missing data (complete survey)
- Variation in survey response rates
- Deprivation correlation (more deprived = longer waits)

**Data generation**: Synthetic data generated using R (`set.seed(42)` for reproducibility). The data uses log-normal distribution to create realistic right skew. Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

**Purpose**: Load and prepare mental health wait times data with realistic skew.

**Key tasks**:
1. Generate synthetic data with skewed distribution
2. Check data structure
3. Explore distribution characteristics

---

::: {.panel-tabset}

### R

```{r ex3-generate-r}
#| message: false
#| warning: false

library(ggplot2)

# Import mental health wait times data
mh_data <- read.csv("data/mental_health_wait_times.csv", stringsAsFactors = FALSE)

# Display first few rows
head(mh_data)

# Summary
summary(mh_data$wait_time_days)
```

### Python

```{python ex3-generate-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import stats

# Import mental health wait times data
mh_data = pd.read_csv("data/mental_health_wait_times.csv")

# Display first few rows
print(mh_data.head())

# Summary
print(mh_data['wait_time_days'].describe())
```

:::

**Initial observations**:
- Dataset has 150 mental health services
- Wait times range from 30 to 215 days
- Mean (79.8 days) > Median (71.5 days) suggests positive skew

---

## Step 2: Assess Distribution

**Purpose**: Determine if data is suitable for standard z-scoring.

---

### Visualize Distribution

::: {.panel-tabset}

### R

```{r ex3-hist-r}
#| fig-width: 10
#| fig-height: 6

ggplot(mh_data, aes(x = wait_time_days)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(mh_data$wait_time_days), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = median(mh_data$wait_time_days), 
             color = "blue", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Mental Health Wait Times",
    subtitle = paste("Mean =", round(mean(mh_data$wait_time_days), 1), 
                    "days | Median =", round(median(mh_data$wait_time_days), 1), "days"),
    x = "Wait Time (days)",
    y = "Number of Services"
  ) +
  theme_minimal()
```

### Python

```{python ex3-hist-py}
plt.figure(figsize=(10, 6))
plt.hist(mh_data['wait_time_days'], bins=30, color='steelblue', edgecolor='white')
plt.axvline(mh_data['wait_time_days'].mean(), color='red', linestyle='--', linewidth=2, label='Mean')
plt.axvline(mh_data['wait_time_days'].median(), color='blue', linestyle='--', linewidth=2, label='Median')
plt.xlabel('Wait Time (days)')
plt.ylabel('Number of Services')
plt.title(f"Distribution of Mental Health Wait Times\nMean = {mh_data['wait_time_days'].mean():.1f} days | Median = {mh_data['wait_time_days'].median():.1f} days")
plt.legend()
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Observations**:
- Clear positive skew (right tail)
- Mean > Median (indicates skewness)
- Most services cluster 60-90 days
- Long tail extends to 200+ days

---

### Check Skewness Statistically

::: {.panel-tabset}

### R

```{r ex3-skew-r}
library(e1071)

skewness_value <- skewness(mh_data$wait_time_days)
cat("Skewness:", round(skewness_value, 2), "\n")
cat("Interpretation:", 
    ifelse(abs(skewness_value) < 0.5, "Approximately symmetric",
    ifelse(skewness_value > 1, "Substantially positively skewed",
    ifelse(skewness_value < -1, "Substantially negatively skewed",
           "Moderately skewed"))), "\n")

# Q-Q plot
ggplot(mh_data, aes(sample = wait_time_days)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Q-Q Plot: Wait Times vs Normal Distribution",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal()
```

### Python

```{python ex3-skew-py}
from scipy.stats import skew

skewness_value = skew(mh_data['wait_time_days'])
print(f"Skewness: {skewness_value:.2f}")

if abs(skewness_value) < 0.5:
    interpretation = "Approximately symmetric"
elif skewness_value > 1:
    interpretation = "Substantially positively skewed"
elif skewness_value < -1:
    interpretation = "Substantially negatively skewed"
else:
    interpretation = "Moderately skewed"
    
print(f"Interpretation: {interpretation}")

# Q-Q plot
fig, ax = plt.subplots(figsize=(8, 8))
stats.probplot(mh_data['wait_time_days'], dist="norm", plot=ax)
ax.set_title("Q-Q Plot: Wait Times vs Normal Distribution")
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Assessment**: 
- Skewness > 1 indicates substantial positive skew
- Q-Q plot shows departure from normality in right tail
- Standard z-scoring assumptions are violated

**Decision**: Compare three approaches:
1. Raw z-scores (despite skewness)
2. Log-transformed z-scores
3. Rank-based approach (percentiles)

---

## Step 3: Approach 1 - Raw Z-scores

**Purpose**: Calculate standard z-scores despite skewness to see the impact.

---

::: {.panel-tabset}

### R

```{r ex3-raw-z-r}
# Calculate raw z-scores
mh_data$z_score_raw <- scale(mh_data$wait_time_days)[,1]

# Assign bands
mh_data$band_raw <- cut(mh_data$z_score_raw,
                        breaks = c(-Inf, -2, -1, 1, 2, Inf),
                        labels = c("Much Better", "Better", "As Expected", 
                                  "Worse", "Much Worse"))

# Summary
table(mh_data$band_raw)

# Flag services in worst band
flagged_raw <- mh_data[mh_data$band_raw == "Much Worse", ]
cat("\nServices flagged (raw z-scores):", nrow(flagged_raw), "\n")
```

### Python

```{python ex3-raw-z-py}
# Calculate raw z-scores
mh_data['z_score_raw'] = (mh_data['wait_time_days'] - mh_data['wait_time_days'].mean()) / mh_data['wait_time_days'].std()

# Assign bands
mh_data['band_raw'] = pd.cut(mh_data['z_score_raw'],
                              bins=[-np.inf, -2, -1, 1, 2, np.inf],
                              labels=["Much Better", "Better", "As Expected", 
                                     "Worse", "Much Worse"])

# Summary
print(mh_data['band_raw'].value_counts().sort_index())

# Flag services in worst band
flagged_raw = mh_data[mh_data['band_raw'] == "Much Worse"]
print(f"\nServices flagged (raw z-scores): {len(flagged_raw)}")
```

:::

---

## Step 4: Approach 2 - Log-Transformed Z-scores

**Purpose**: Apply log transformation to reduce skew, then calculate z-scores.

---

::: {.panel-tabset}

### R

```{r ex3-log-z-r}
# Log transform wait times
mh_data$log_wait_time <- log(mh_data$wait_time_days)

# Calculate z-scores on log scale
mh_data$z_score_log <- scale(mh_data$log_wait_time)[,1]

# Assign bands
mh_data$band_log <- cut(mh_data$z_score_log,
                        breaks = c(-Inf, -2, -1, 1, 2, Inf),
                        labels = c("Much Better", "Better", "As Expected", 
                                  "Worse", "Much Worse"))

# Summary
table(mh_data$band_log)

# Flag services in worst band
flagged_log <- mh_data[mh_data$band_log == "Much Worse", ]
cat("\nServices flagged (log z-scores):", nrow(flagged_log), "\n")

# Check distribution of log-transformed data
ggplot(mh_data, aes(x = log_wait_time)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Log-Transformed Wait Times",
    subtitle = "More symmetric after transformation",
    x = "Log(Wait Time)",
    y = "Number of Services"
  ) +
  theme_minimal()
```

### Python

```{python ex3-log-z-py}
# Log transform wait times
mh_data['log_wait_time'] = np.log(mh_data['wait_time_days'])

# Calculate z-scores on log scale
mh_data['z_score_log'] = (mh_data['log_wait_time'] - mh_data['log_wait_time'].mean()) / mh_data['log_wait_time'].std()

# Assign bands
mh_data['band_log'] = pd.cut(mh_data['z_score_log'],
                              bins=[-np.inf, -2, -1, 1, 2, np.inf],
                              labels=["Much Better", "Better", "As Expected", 
                                     "Worse", "Much Worse"])

# Summary
print(mh_data['band_log'].value_counts().sort_index())

# Flag services in worst band
flagged_log = mh_data[mh_data['band_log'] == "Much Worse"]
print(f"\nServices flagged (log z-scores): {len(flagged_log)}")

# Check distribution of log-transformed data
plt.figure(figsize=(10, 6))
plt.hist(mh_data['log_wait_time'], bins=30, color='steelblue', edgecolor='white')
plt.xlabel('Log(Wait Time)')
plt.ylabel('Number of Services')
plt.title('Distribution of Log-Transformed Wait Times\nMore symmetric after transformation')
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Observations**:
- Log transformation reduces skewness
- Distribution closer to normal
- Different services may be flagged compared to raw approach

---

## Step 5: Compare Approaches

**Purpose**: Assess sensitivity of results to methodological choice.

---

::: {.panel-tabset}

### R

```{r ex3-compare-r}
# Compare flagged services
raw_codes <- mh_data$service_code[mh_data$band_raw == "Much Worse"]
log_codes <- mh_data$service_code[mh_data$band_log == "Much Worse"]

# Agreement
agreement <- length(intersect(raw_codes, log_codes))
raw_only <- length(setdiff(raw_codes, log_codes))
log_only <- length(setdiff(log_codes, raw_codes))

cat("Services flagged by both methods:", agreement, "\n")
cat("Flagged by raw z-scores only:", raw_only, "\n")
cat("Flagged by log z-scores only:", log_only, "\n")
cat("Agreement rate:", round(100 * agreement / max(length(raw_codes), length(log_codes)), 1), "%\n")

# Scatter plot comparing approaches
ggplot(mh_data, aes(x = z_score_raw, y = z_score_log)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 2, color = "red", linetype = "dashed") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed") +
  labs(
    title = "Comparison of Raw vs Log-Transformed Z-scores",
    subtitle = "Dashed lines show flagging threshold (z > 2)",
    x = "Raw Z-score",
    y = "Log-Transformed Z-score"
  ) +
  theme_minimal()
```

### Python

```{python ex3-compare-py}
# Compare flagged services
raw_codes = set(mh_data[mh_data['band_raw'] == "Much Worse"]['service_code'])
log_codes = set(mh_data[mh_data['band_log'] == "Much Worse"]['service_code'])

# Agreement
agreement = len(raw_codes & log_codes)
raw_only = len(raw_codes - log_codes)
log_only = len(log_codes - raw_codes)

print(f"Services flagged by both methods: {agreement}")
print(f"Flagged by raw z-scores only: {raw_only}")
print(f"Flagged by log z-scores only: {log_only}")
print(f"Agreement rate: {100 * agreement / max(len(raw_codes), len(log_codes)):.1f}%")

# Scatter plot comparing approaches
plt.figure(figsize=(10, 8))
plt.scatter(mh_data['z_score_raw'], mh_data['z_score_log'], alpha=0.6)
plt.axhline(y=2, color='red', linestyle='--', label='Flagging threshold')
plt.axvline(x=2, color='red', linestyle='--')
plt.xlabel('Raw Z-score')
plt.ylabel('Log-Transformed Z-score')
plt.title('Comparison of Raw vs Log-Transformed Z-scores\nDashed lines show flagging threshold (z > 2)')
plt.grid(True, alpha=0.3)
plt.legend()
plt.tight_layout()
plt.show()
```

:::

---

## Step 6: Recommendation

**Purpose**: Document analytical decision and rationale.

---

**Decision**: Use log-transformed z-scores for final analysis.

**Rationale**:
1. **Assumption validity**: Log transformation brings data closer to normality
2. **Statistical appropriateness**: Z-scoring assumes normal distribution
3. **Reduced sensitivity to outliers**: Extreme values have less influence
4. **Interpretability**: Can back-transform for reporting

**Caveats**:
- Log transformation changes the metric (multiplicative vs additive differences)
- Services near threshold may differ between methods
- Document both approaches in QA log

**Communication**: Report wait times in original scale (days) but use log-transformed z-scores for statistical comparison.

---

## Step 7: Final Results

::: {.panel-tabset}

### R

```{r ex3-final-r}
# Services requiring follow-up (log z-score approach)
flagged_final <- mh_data[mh_data$band_log %in% c("Worse", "Much Worse"), 
                         c("service_code", "service_name", "wait_time_days", 
                           "z_score_log", "band_log")]

flagged_final <- flagged_final[order(-flagged_final$z_score_log), ]

cat("Services flagged for follow-up:", nrow(flagged_final), "\n")
cat("Percentage of total:", round(100 * nrow(flagged_final) / nrow(mh_data), 1), "%\n\n")

# Top 10
head(flagged_final, 10)
```

### Python

```{python ex3-final-py}
# Services requiring follow-up (log z-score approach)
flagged_final = mh_data[mh_data['band_log'].isin(["Worse", "Much Worse"])][
    ['service_code', 'service_name', 'wait_time_days', 'z_score_log', 'band_log']
].sort_values('z_score_log', ascending=False)

print(f"Services flagged for follow-up: {len(flagged_final)}")
print(f"Percentage of total: {100 * len(flagged_final) / len(mh_data):.1f}%\n")

# Top 10
print(flagged_final.head(10))
```

:::

---

**Analysis Complete** ✅

**Key Takeaways**:
1. Always check distribution assumptions before applying z-scoring
2. Skewed data can be handled with transformation
3. Compare multiple approaches for sensitivity
4. Document methodological decisions
5. Report results in interpretable units

---
