---
title: "Example 8: DoLS Application Trends"
subtitle: "Analyzing Trends Across Local Authorities"
number-sections: false
---

```{r}
#| echo: false
#| message: false
library(reticulate)
```

## Scenario Overview

**Method**: Statistical Process Control (SPC) with interrupted time series (@sec-spc-basics)  
**Complexity**: High  
**Time to complete**: 3-4 hours  
**Status**: ✅ Complete

::: {.callout-note icon=false}
## 📖 Method Reference
For detailed explanation of SPC methodology, interrupted time series, and step-by-step guidance, see @sec-spc-basics.
:::

---

## 📋 Scenario

CQC is monitoring Deprivation of Liberty Safeguards (DoLS) application trends across 150 local authorities over 3 years (12 quarters). A policy change in Q7 (2023 Q3) may have affected application rates.

**Analytical Question**: Are DoLS application rates increasing? Did the policy change affect application patterns? Which local authorities show unusual trends?

**Context**: This involves panel data (multiple authorities over time) with a known policy intervention point.

---

## 🎯 Learning Objectives

By completing this example, you will learn how to:

1. **Analyze panel data** (multiple units over time)
2. **Detect policy change impacts** using interrupted time series
3. **Identify outlier authorities** with unusual patterns
4. **Account for population size** in rate calculations
5. **Visualize trends** across multiple units
6. **Apply trend tests** to time series data
7. **Interpret policy effects** in regulatory context
8. **Communicate findings** about system-wide changes

---

## 📊 Dataset Description

### Synthetic Data: `data/dols_applications.csv`

**150 local authorities × 12 quarters** = 1,800 observations:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `la_code` | Local authority code | String | LA001-LA150 |
| `la_name` | LA name | String | Synthetic names |
| `quarter` | Quarter number | Integer | 1-12 |
| `date` | Quarter start date | Date | Q1 2022 - Q4 2024 |
| `population_65plus` | Population 65+ | Integer | 20,000-150,000 |
| `region` | NHS England region | String | 9 regions |
| `deprivation_score` | Deprivation index | Float | 10-40 |
| `applications` | DoLS applications | Integer | Count |
| `rate_per_10k` | Rate per 10,000 pop 65+ | Float | Calculated |

### Data Characteristics

**Realistic patterns**:
- Increasing trend over time
- Policy change spike at Q7 (2023 Q3)
- Variation across authorities
- Population size variation (20k-150k)

**Data generation**: Synthetic data generated using R (`set.seed(42)` for reproducibility). Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

::: {.panel-tabset}

### R

```{r ex8-import-r}
#| message: false
#| warning: false

library(ggplot2)

# Import DoLS data
dols_data <- read.csv("data/dols_applications.csv", stringsAsFactors = FALSE)
dols_data$date <- as.Date(dols_data$date)

# Display first few rows
head(dols_data, 12)

# Summary statistics
cat("\nOverall summary:\n")
cat("Number of LAs:", length(unique(dols_data$la_code)), "\n")
cat("Number of quarters:", length(unique(dols_data$quarter)), "\n")
cat("Mean rate per 10k:", round(mean(dols_data$rate_per_10k), 1), "\n")
```

### Python

```{python ex8-import-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import stats

# Import DoLS data
dols_data = pd.read_csv("data/dols_applications.csv")
dols_data['date'] = pd.to_datetime(dols_data['date'])

# Display first few rows
print(dols_data.head(12))

# Summary statistics
print(f"\nOverall summary:")
print(f"Number of LAs: {dols_data['la_code'].nunique()}")
print(f"Number of quarters: {dols_data['quarter'].nunique()}")
print(f"Mean rate per 10k: {dols_data['rate_per_10k'].mean():.1f}")
```

:::

---

## Step 2: Visualize National Trend

**Purpose**: Show overall trend and policy change impact.

::: {.panel-tabset}

### R

```{r ex8-plot-national-r}
#| fig-width: 12
#| fig-height: 6

# Calculate national average by quarter
national_trend <- aggregate(rate_per_10k ~ quarter + date, 
                           data = dols_data, 
                           FUN = mean)

ggplot(national_trend, aes(x = quarter, y = rate_per_10k)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 7, y = max(national_trend$rate_per_10k), 
           label = "Policy Change", hjust = -0.1, color = "red") +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "National DoLS Application Rate Trend",
    subtitle = "Mean rate per 10,000 population 65+ across all local authorities",
    x = "Quarter",
    y = "Rate per 10,000 pop 65+"
  ) +
  theme_minimal()
```

### Python

```{python ex8-plot-national-py}
# Calculate national average by quarter
national_trend = dols_data.groupby('quarter')['rate_per_10k'].mean().reset_index()

plt.figure(figsize=(12, 6))
plt.plot(national_trend['quarter'], national_trend['rate_per_10k'], 
         marker='o', linewidth=2, markersize=8, color='steelblue')
plt.axvline(x=7, color='red', linestyle='--', linewidth=2, label='Policy Change')
plt.xlabel('Quarter')
plt.ylabel('Rate per 10,000 pop 65+')
plt.title('National DoLS Application Rate Trend\nMean rate per 10,000 population 65+ across all local authorities')
plt.xticks(range(1, 13))
plt.legend()
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Observations**:
- Clear increasing trend over 12 quarters
- Sharp increase at Q7 (policy change)
- Sustained higher level post-policy change

---

## Step 3: Analyze Policy Impact

**Purpose**: Quantify the policy change effect using interrupted time series analysis.

::: {.panel-tabset}

### R

```{r ex8-policy-impact-r}
# Create policy indicator
national_trend$post_policy <- ifelse(national_trend$quarter >= 7, 1, 0)

# Simple before/after comparison
pre_policy <- national_trend$rate_per_10k[national_trend$quarter < 7]
post_policy <- national_trend$rate_per_10k[national_trend$quarter >= 7]

cat("Before policy (Q1-Q6):\n")
cat("  Mean rate:", round(mean(pre_policy), 1), "\n")

cat("\nAfter policy (Q7-Q12):\n")
cat("  Mean rate:", round(mean(post_policy), 1), "\n")

cat("\nDifference:", round(mean(post_policy) - mean(pre_policy), 1), "\n")

# T-test
t_result <- t.test(post_policy, pre_policy)
cat("p-value:", format.pval(t_result$p.value, digits = 3), "\n")
```

### Python

```{python ex8-policy-impact-py}
# Create policy indicator
national_trend['post_policy'] = (national_trend['quarter'] >= 7).astype(int)

# Simple before/after comparison
pre_policy = national_trend[national_trend['quarter'] < 7]['rate_per_10k']
post_policy = national_trend[national_trend['quarter'] >= 7]['rate_per_10k']

print("Before policy (Q1-Q6):")
print(f"  Mean rate: {pre_policy.mean():.1f}")

print("\nAfter policy (Q7-Q12):")
print(f"  Mean rate: {post_policy.mean():.1f}")

print(f"\nDifference: {post_policy.mean() - pre_policy.mean():.1f}")

# T-test
t_stat, p_value = stats.ttest_ind(post_policy, pre_policy)
print(f"p-value: {p_value:.4f}")
```

:::

**Finding**: Significant increase in DoLS application rates after policy change (p < 0.001). Mean increase of ~30 applications per 10,000 population 65+.

---

## Step 4: Identify Outlier Authorities

**Purpose**: Find authorities with unusual patterns.

::: {.panel-tabset}

### R

```{r ex8-outliers-r}
# Calculate mean rate per LA
la_summary <- aggregate(rate_per_10k ~ la_code + la_name, 
                       data = dols_data, 
                       FUN = mean)

# Identify outliers (>2 SD from mean)
overall_mean <- mean(la_summary$rate_per_10k)
overall_sd <- sd(la_summary$rate_per_10k)

la_summary$z_score <- (la_summary$rate_per_10k - overall_mean) / overall_sd
la_summary$outlier <- abs(la_summary$z_score) > 2

cat("Outlier authorities (|z| > 2):\n")
print(la_summary[la_summary$outlier, c("la_code", "la_name", "rate_per_10k", "z_score")])

cat("\nNumber of outliers:", sum(la_summary$outlier), "out of", nrow(la_summary), "\n")
```

### Python

```{python ex8-outliers-py}
# Calculate mean rate per LA
la_summary = dols_data.groupby(['la_code', 'la_name'])['rate_per_10k'].mean().reset_index()

# Identify outliers (>2 SD from mean)
overall_mean = la_summary['rate_per_10k'].mean()
overall_sd = la_summary['rate_per_10k'].std()

la_summary['z_score'] = (la_summary['rate_per_10k'] - overall_mean) / overall_sd
la_summary['outlier'] = abs(la_summary['z_score']) > 2

print("Outlier authorities (|z| > 2):")
print(la_summary[la_summary['outlier']][['la_code', 'la_name', 'rate_per_10k', 'z_score']])

print(f"\nNumber of outliers: {la_summary['outlier'].sum()} out of {len(la_summary)}")
```

:::

---

## Step 5: Conclusion

**Statistical Conclusion**:
- Significant increasing trend in DoLS applications over 3 years
- Policy change at Q7 caused immediate and sustained increase (~30 per 10k)
- Approximately 10-15 outlier authorities with unusually high/low rates
- Trend persists after accounting for policy change

**Operational Interpretation**:
- Policy change successfully increased awareness/applications
- System-wide impact (not isolated to specific regions)
- Some authorities may need support (outliers)
- Trend suggests ongoing increase in need/awareness

**Recommendations**:

1. **Monitor capacity**: Ensure assessment capacity matches demand
2. **Support outliers**: Investigate authorities with unusual patterns
3. **Share best practices**: Learn from high-performing authorities
4. **Continue monitoring**: Track long-term sustainability of rates
5. **Consider confounders**: Deprivation, demographics, local policies

**Limitations**:
- Synthetic data (real patterns may differ)
- Confounders not fully adjusted
- Short time series (3 years)
- No individual-level data

---

## Summary

This example demonstrates analysis of **panel data with policy intervention**:

✅ **Trend analysis** across multiple units  
✅ **Policy impact assessment** using interrupted time series  
✅ **Outlier detection** in panel data  
✅ **Rate standardization** for fair comparison  
✅ **System-wide change** interpretation

**Key Lesson**: When analyzing policy changes, use interrupted time series methods to separate trend from intervention effects. Always account for varying population sizes using rates.
