---
title: "Example 7: Infection Rates in Care Homes"
subtitle: "Comparing Rates with Varying Denominators"
number-sections: false
---

```{r}
#| echo: false
#| message: false
library(reticulate)
use_python("C:/Users/fede/anaconda3/python.exe")
```

## Scenario Overview

**Method**: Multi-group rate comparison (requires Guild consultation for ANOVA/chi-squared)  
**Complexity**: High  
**Time to complete**: 3-4 hours  
**Status**: ✅ Complete

::: {.callout-note icon=false}
## 📖 Method Reference
This example demonstrates comparing rates across multiple groups. For background on rate calculation and standardization, see @sec-z-scoring. For multi-group comparisons beyond two groups, consult the Quantitative Guild.
:::

---

## 📋 Scenario

CQC is analyzing infection rates across 80 care homes using Vivaldi study data. Care homes vary in size (20-100 residents) and quality ratings.

**Analytical Question**: Do infection rates differ by CQC rating? How should we account for varying home sizes?

**Context**: This involves count data (infections) with varying denominators (residents). Simple comparisons may be misleading if not properly adjusted for exposure.

---

## 🎯 Learning Objectives

By completing this example, you will learn how to:

1. **Work with count data** and rates
2. **Account for varying denominators** (exposure adjustment)
3. **Compare rates across groups** (CQC ratings)
4. **Handle small numbers** in some homes
5. **Interpret rate vs count** differences
6. **Apply appropriate statistical tests** for count data
7. **Visualize rates** with confidence intervals
8. **Communicate findings** about quality and safety

---

## 📊 Dataset Description

### Synthetic Data: `data/infection_rates.csv`

**80 care homes** with infection data:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `home_id` | Care home identifier | String | CH001-CH080 |
| `home_name` | Care home name | String | Synthetic names |
| `residents` | Number of residents | Integer | 20-100 |
| `staff_ratio` | Staff-to-resident ratio | Float | 0.3-0.7 |
| `cqc_rating` | CQC quality rating | String | Outstanding/Good/RI/Inadequate |
| `infections` | Number of infections | Integer | Count data |
| `infection_rate` | Rate per 100 residents | Float | Calculated |

### Data Characteristics

**Realistic patterns**:
- Varying home sizes (20-100 residents)
- Most homes rated "Good" (70%)
- Infection rates related to quality rating
- Small numbers in some homes
- Poisson-distributed counts

**Data generation**: Synthetic data generated using R (`set.seed(42)` for reproducibility). Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

::: {.panel-tabset}

### R

```{r ex7-import-r}
#| message: false
#| warning: false

library(ggplot2)

# Import infection data
infection_data <- read.csv("data/infection_rates.csv", stringsAsFactors = FALSE)

# Display first few rows
head(infection_data)

# Summary by rating
cat("\nSummary by CQC Rating:\n")
aggregate(cbind(residents, infections, infection_rate) ~ cqc_rating, 
          data = infection_data, 
          FUN = function(x) c(n = length(x), mean = round(mean(x), 2)))
```

### Python

```{python ex7-import-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import stats

# Import infection data
infection_data = pd.read_csv("data/infection_rates.csv")

# Display first few rows
print(infection_data.head())

# Summary by rating
print("\nSummary by CQC Rating:")
print(infection_data.groupby('cqc_rating')[['residents', 'infections', 'infection_rate']].agg(['count', 'mean']))
```

:::

---

## Step 2: Visualize Rates by Quality

::: {.panel-tabset}

### R

```{r ex7-plot-r}
#| fig-width: 10
#| fig-height: 6

# Order ratings
infection_data$cqc_rating <- factor(infection_data$cqc_rating, 
                                    levels = c("Outstanding", "Good", 
                                              "Requires Improvement", "Inadequate"))

ggplot(infection_data, aes(x = cqc_rating, y = infection_rate, fill = cqc_rating)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("Outstanding" = "#2E7D32", 
                               "Good" = "#66BB6A",
                               "Requires Improvement" = "#FFA726",
                               "Inadequate" = "#EF5350")) +
  labs(
    title = "Infection Rates by CQC Rating",
    subtitle = "Rate per 100 residents",
    x = "CQC Rating",
    y = "Infection Rate (per 100 residents)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Python

```{python ex7-plot-py}
# Order ratings
rating_order = ["Outstanding", "Good", "Requires Improvement", "Inadequate"]
infection_data['cqc_rating'] = pd.Categorical(infection_data['cqc_rating'], 
                                               categories=rating_order, 
                                               ordered=True)

plt.figure(figsize=(10, 6))
infection_data.boxplot(column='infection_rate', by='cqc_rating', 
                       grid=False, patch_artist=True)
plt.suptitle('')
plt.title('Infection Rates by CQC Rating\nRate per 100 residents')
plt.xlabel('CQC Rating')
plt.ylabel('Infection Rate (per 100 residents)')
plt.xticks(rotation=45, ha='right')
plt.tight_layout()
plt.show()
```

:::

**Observations**:
- Clear gradient: Outstanding < Good < RI < Inadequate
- Variation within each rating category
- Some overlap between adjacent categories

---

## Step 3: Statistical Comparison

**Purpose**: Test if infection rates differ significantly by CQC rating.

::: {.panel-tabset}

### R

```{r ex7-test-r}
# ANOVA on rates
anova_result <- aov(infection_rate ~ cqc_rating, data = infection_data)
summary(anova_result)

# Pairwise comparisons
pairwise_result <- pairwise.t.test(infection_data$infection_rate, 
                                   infection_data$cqc_rating,
                                   p.adjust.method = "bonferroni")
print(pairwise_result)

# Mean rates by rating
cat("\nMean infection rates by CQC rating:\n")
tapply(infection_data$infection_rate, infection_data$cqc_rating, mean)
```

### Python

```{python ex7-test-py}
# ANOVA on rates
from scipy.stats import f_oneway

groups = [infection_data[infection_data['cqc_rating'] == rating]['infection_rate'].values 
          for rating in rating_order]

f_stat, p_value = f_oneway(*groups)
print(f"ANOVA F-statistic: {f_stat:.2f}")
print(f"p-value: {p_value:.4f}")

# Mean rates by rating
print("\nMean infection rates by CQC rating:")
print(infection_data.groupby('cqc_rating')['infection_rate'].mean())
```

:::

**Finding**: Significant differences in infection rates by CQC rating (p < 0.001).

---

## Step 4: Conclusion

**Statistical Conclusion**:
- Significant association between CQC rating and infection rates
- Outstanding homes: ~2% infection rate
- Good homes: ~3% infection rate  
- Requires Improvement: ~4% infection rate
- Inadequate: ~5% infection rate

**Practical Significance**:
- Quality rating is a strong predictor of infection control
- Difference between Outstanding and Inadequate: ~3 percentage points
- In a 50-resident home: ~1.5 fewer infections per period

**Recommendations**:
1. **Target support** to Requires Improvement and Inadequate homes
2. **Share best practices** from Outstanding homes
3. **Monitor infection rates** as quality indicator
4. **Consider confounders** (resident case-mix, staffing levels)

**Limitations**:
- Cross-sectional data (no causality)
- Resident case-mix not fully adjusted
- Small sample sizes in some rating categories
- Reporting bias possible

---

## Summary

This example demonstrates analysis of **count data with varying denominators**:

✅ **Rate calculation** and interpretation  
✅ **Comparison across groups** with different exposures  
✅ **Appropriate statistical tests** for rates  
✅ **Visualization** of rates with context  
✅ **Quality-safety relationship** interpretation

**Key Lesson**: When comparing counts across units with different sizes, always use rates (per 100, per 1000, etc.) to ensure fair comparison.
