---
title: "Example 6: Ambulance Handover Delays"
subtitle: "SPC with Trend Analysis and Seasonality"
number-sections: false
---

```{r}
#| echo: false
#| message: false
library(reticulate)
```

## Scenario Overview

**Method**: Statistical Process Control (SPC) (@sec-spc-basics)  
**Complexity**: Medium-High  
**Time to complete**: 2-3 hours  
**Status**: ✅ Complete

::: {.callout-note icon=false}
## 📖 Method Reference
For detailed explanation of SPC methodology, control charts with trends, and step-by-step guidance, see @sec-spc-basics.
:::

---

## 📋 Scenario

A hospital trust is monitoring ambulance handover delays (>30 minutes) over 18 months. National guidance states handovers should occur within 15 minutes, with 30 minutes as the maximum acceptable threshold.

**Analytical Question**: Are handover delays increasing over time? Are there special causes requiring investigation?

**Context**: This is operational performance monitoring with clear system pressures. The trust needs to understand whether delays are part of common cause variation or indicate special causes requiring intervention.

---

## 🎯 Learning Objectives

By completing this example, you will learn how to:

1. **Apply SPC to trending data** with increasing baseline
2. **Identify and handle seasonality** (winter pressures)
3. **Detect special causes** in the presence of trends
4. **Distinguish trends from shifts** in process performance
5. **Use trend rules** in SPC (Rule 5: runs above/below centre)
6. **Interpret SPC in operational context** (system capacity issues)
7. **Communicate findings** about deteriorating performance
8. **Document decisions** about when to recalculate limits

---

## 📊 Dataset Description

### Synthetic Data: `data/ambulance_handovers.csv`

**18 months** of handover delay data:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `month` | Month number | Integer | 1-18 |
| `date` | Month start date | Date | Jan 2023 - Jun 2024 |
| `delays_over_30min` | Number of handovers >30 min | Integer | 100-310 per month |

### Data Characteristics

**Realistic patterns**:
- Increasing trend (~5 delays/month increase)
- Winter peaks (Dec-Feb): +40 delays
- Special causes: IT failure (Month 8), flu outbreak (Month 15)
- Natural variation throughout

**Data generation**: Synthetic data generated using R (`set.seed(42)` for reproducibility). Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

**Purpose**: Load and visualize the time series to understand patterns.

---

### Import Data

::: {.panel-tabset}

### R

```{r ex6-import-r}
#| message: false
#| warning: false

library(ggplot2)

# Import handover data
handover_data <- read.csv("data/ambulance_handovers.csv", stringsAsFactors = FALSE)
handover_data$date <- as.Date(handover_data$date)

# Display data
print(handover_data)

# Summary
cat("\nMean delays:", round(mean(handover_data$delays_over_30min), 1), "\n")
cat("Range:", min(handover_data$delays_over_30min), "to", max(handover_data$delays_over_30min), "\n")
```

### Python

```{python ex6-import-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import stats

# Import handover data
handover_data = pd.read_csv("data/ambulance_handovers.csv")
handover_data['date'] = pd.to_datetime(handover_data['date'])

# Display data
print(handover_data)

# Summary
print(f"\nMean delays: {handover_data['delays_over_30min'].mean():.1f}")
print(f"Range: {handover_data['delays_over_30min'].min()} to {handover_data['delays_over_30min'].max()}")
```

:::

**Initial observations**:
- 18 months of data
- Delays range from ~140 to ~310 per month
- Visual inspection needed to identify patterns

---

## Step 2: Visualize Time Series

**Purpose**: Identify trends, seasonality, and potential special causes.

---

::: {.panel-tabset}

### R

```{r ex6-plot-ts-r}
ggplot(handover_data, aes(x = month, y = delays_over_30min)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(breaks = 1:18) +
  labs(
    title = "Ambulance Handover Delays Over Time",
    subtitle = "Number of handovers >30 minutes per month",
    x = "Month",
    y = "Delays >30 min"
  ) +
  theme_minimal()
```

### Python

```{python ex6-plot-ts-py}
plt.figure(figsize=(12, 6))
plt.plot(handover_data['month'], handover_data['delays_over_30min'], 
         marker='o', linewidth=2, markersize=8, color='steelblue')
plt.xlabel('Month')
plt.ylabel('Delays >30 min')
plt.title('Ambulance Handover Delays Over Time\nNumber of handovers >30 minutes per month')
plt.xticks(range(1, 19))
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Observations**:
- Clear increasing trend over 18 months
- Peaks visible at months 8 and 15
- Possible seasonal pattern (winter months)
- Baseline appears unstable (not suitable for standard SPC)

---

## Step 3: Calculate Control Limits

**Purpose**: Apply SPC despite trending data to identify special causes.

**Note**: With trending data, control limits may not be appropriate for the full series. We calculate them to demonstrate the issue.

---

::: {.panel-tabset}

### R

```{r ex6-calc-limits-r}
# Calculate centre line (mean)
centre_line <- mean(handover_data$delays_over_30min)

# Calculate moving range
mr <- abs(diff(handover_data$delays_over_30min))
mean_mr <- mean(mr)

# Calculate control limits
ucl <- centre_line + 2.66 * mean_mr
lcl <- centre_line - 2.66 * mean_mr

cat("Control Limits (full series):\n")
cat("  Centre Line:", round(centre_line, 1), "\n")
cat("  UCL:", round(ucl, 1), "\n")
cat("  LCL:", round(lcl, 1), "\n")

# Identify points outside limits
handover_data$outside_limits <- handover_data$delays_over_30min > ucl | 
                                 handover_data$delays_over_30min < lcl

cat("\nPoints outside limits:\n")
print(handover_data[handover_data$outside_limits, c("month", "date", "delays_over_30min")])
```

### Python

```{python ex6-calc-limits-py}
# Calculate centre line
centre_line = handover_data['delays_over_30min'].mean()

# Calculate moving range
mr = handover_data['delays_over_30min'].diff().abs().dropna()
mean_mr = mr.mean()

# Calculate control limits
ucl = centre_line + 2.66 * mean_mr
lcl = centre_line - 2.66 * mean_mr

print("Control Limits (full series):")
print(f"  Centre Line: {centre_line:.1f}")
print(f"  UCL: {ucl:.1f}")
print(f"  LCL: {lcl:.1f}")

# Identify points outside limits
handover_data['outside_limits'] = (handover_data['delays_over_30min'] > ucl) | \
                                   (handover_data['delays_over_30min'] < lcl)

print("\nPoints outside limits:")
print(handover_data[handover_data['outside_limits']][['month', 'date', 'delays_over_30min']])
```

:::

**Finding**: Months 8 and 15 are above UCL - special causes detected.

---

## Step 4: Create SPC Chart

**Purpose**: Visualize data with control limits and special causes.

---

::: {.panel-tabset}

### R

```{r ex6-spc-chart-r}
#| fig-width: 12
#| fig-height: 7

ggplot(handover_data, aes(x = month, y = delays_over_30min)) +
  geom_hline(yintercept = centre_line, color = "blue", linewidth = 1, linetype = "solid") +
  geom_hline(yintercept = ucl, color = "red", linewidth = 1, linetype = "dashed") +
  geom_hline(yintercept = lcl, color = "red", linewidth = 1, linetype = "dashed") +
  geom_line(color = "gray50", linewidth = 0.5) +
  geom_point(aes(color = outside_limits), size = 4) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "red"),
                     labels = c("Within limits", "Outside limits")) +
  scale_x_continuous(breaks = 1:18) +
  annotate("text", x = 8, y = handover_data$delays_over_30min[8] + 15, 
           label = "IT failure", size = 3) +
  annotate("text", x = 15, y = handover_data$delays_over_30min[15] + 15, 
           label = "Flu outbreak", size = 3) +
  labs(
    title = "SPC Chart: Ambulance Handover Delays",
    subtitle = "Note: Trending data makes standard SPC limits less meaningful",
    x = "Month",
    y = "Delays >30 min",
    color = "Point Status"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Python

```{python ex6-spc-chart-py}
plt.figure(figsize=(14, 7))

# Plot centre line and limits
plt.axhline(y=centre_line, color='blue', linewidth=2, label='Centre Line')
plt.axhline(y=ucl, color='red', linewidth=2, linestyle='--', label='UCL')
plt.axhline(y=lcl, color='red', linewidth=2, linestyle='--', label='LCL')

# Plot data
normal_points = handover_data[~handover_data['outside_limits']]
special_points = handover_data[handover_data['outside_limits']]

plt.plot(handover_data['month'], handover_data['delays_over_30min'], 
         color='gray', linewidth=1, alpha=0.5)
plt.scatter(normal_points['month'], normal_points['delays_over_30min'], 
           color='steelblue', s=100, label='Within limits', zorder=3)
plt.scatter(special_points['month'], special_points['delays_over_30min'], 
           color='red', s=100, label='Outside limits', zorder=3)

# Annotations
plt.annotate('IT failure', xy=(8, handover_data.loc[7, 'delays_over_30min']), 
            xytext=(8, handover_data.loc[7, 'delays_over_30min'] + 20),
            ha='center', fontsize=9)
plt.annotate('Flu outbreak', xy=(15, handover_data.loc[14, 'delays_over_30min']), 
            xytext=(15, handover_data.loc[14, 'delays_over_30min'] + 20),
            ha='center', fontsize=9)

plt.xlabel('Month')
plt.ylabel('Delays >30 min')
plt.title('SPC Chart: Ambulance Handover Delays\nNote: Trending data makes standard SPC limits less meaningful')
plt.xticks(range(1, 19))
plt.legend(loc='upper left')
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

---

## Step 5: Assess Trend

**Purpose**: Quantify the increasing trend using regression.

---

::: {.panel-tabset}

### R

```{r ex6-trend-r}
# Linear regression
trend_model <- lm(delays_over_30min ~ month, data = handover_data)
summary(trend_model)

# Extract trend
slope <- coef(trend_model)[2]
intercept <- coef(trend_model)[1]

cat("\nTrend Analysis:\n")
cat("  Slope:", round(slope, 2), "delays per month\n")
cat("  Intercept:", round(intercept, 1), "\n")
cat("  R-squared:", round(summary(trend_model)$r.squared, 3), "\n")

# Predict values
handover_data$predicted <- predict(trend_model)
```

### Python

```{python ex6-trend-py}
# Linear regression
from scipy.stats import linregress

slope, intercept, r_value, p_value, std_err = linregress(
    handover_data['month'], 
    handover_data['delays_over_30min']
)

print("\nTrend Analysis:")
print(f"  Slope: {slope:.2f} delays per month")
print(f"  Intercept: {intercept:.1f}")
print(f"  R-squared: {r_value**2:.3f}")
print(f"  p-value: {p_value:.4f}")

# Predict values
handover_data['predicted'] = intercept + slope * handover_data['month']
```

:::

**Finding**: Significant increasing trend of ~5 delays per month (p < 0.001). This confirms the process is not stable.

---

## Step 6: Conclusion and Recommendations

**Statistical Conclusion**:
- Significant increasing trend in handover delays
- Two special causes identified (Months 8 and 15)
- Process is not in statistical control
- Standard SPC limits not appropriate for trending data

**Operational Interpretation**:
- System capacity deteriorating over time
- Winter pressures compound the baseline trend
- Special causes (IT failure, flu outbreak) create additional spikes
- Trend suggests systemic issues, not just variation

**Recommendations**:

1. **Immediate actions**:
   - Investigate root causes of increasing trend
   - Review capacity planning and staffing levels
   - Prepare for winter pressures (seasonal peaks)

2. **Monitoring approach**:
   - Use trend analysis rather than standard SPC
   - Set target trajectory for improvement
   - Monitor deviations from expected trend
   - Recalculate limits if trend stabilizes

3. **System improvements**:
   - Address underlying capacity constraints
   - Improve handover processes
   - Enhance winter resilience planning
   - Implement early warning systems

**Limitations**:
- Only 18 months of data (limited for seasonal analysis)
- Single trust (generalizability unknown)
- External factors (e.g., regional demand) not captured
- Trend may not be linear long-term

---

## Summary

This example demonstrates SPC application to **trending data**, showing:

✅ **Trend detection** using regression analysis  
✅ **Special cause identification** despite trend  
✅ **Limitations of standard SPC** with unstable processes  
✅ **Operational context** interpretation (capacity issues)  
✅ **Alternative monitoring approaches** for trending data

**Key Lesson**: When processes show sustained trends, standard SPC control limits become less meaningful. Focus on trend analysis and investigate root causes of deterioration.
