---
title: "Example 2: Care Home Falls Monitoring"
subtitle: "Statistical Process Control with Intervention"
number-sections: false
---

```{r}
#| include: false
# Setup Python environment
library(reticulate)
use_python("C:/Users/fede/anaconda3/python.exe", required = TRUE)
```

## Scenario Overview

**Method**: Statistical Process Control (SPC) (@sec-spc-basics)  
**Complexity**: Medium  
**Time to complete**: 2-3 hours  
**Status**: ✅ Complete

::: {.callout-note icon=false}
## 📖 Method Reference
For detailed explanation of SPC methodology, control charts, and step-by-step guidance, see @sec-spc-basics.
:::

---

## 📋 Scenario

You are a CQC analyst monitoring falls rates at Oakwood Care Home over 24 months. A falls prevention program was implemented in January 2024 (Month 13). Your task is to determine whether the intervention reduced falls rates and whether the improvement is sustained.

**Analytical Question**: Did the falls prevention program reduce falls rates at Oakwood Care Home?

**Context**: Falls are a key quality indicator for care homes. Oakwood Care Home has been tracking monthly falls for 2 years and implemented a comprehensive falls prevention program including staff training, environmental modifications, and enhanced monitoring.

---

## 🎯 Learning Objectives

By working through this example, you will learn to:

1. **Set up control charts** from scratch using I-chart methodology
2. **Calculate control limits** correctly (centre line, UCL, LCL)
3. **Identify special causes** and decide when to exclude them
4. **Detect intervention effects** using SPC detection rules
5. **Apply detection rules** (Rule 1: outside limits, Rule 4: runs)
6. **Assess process stability** before and after intervention
7. **Evaluate sustainability** of improvements
8. **Communicate SPC results** to non-technical stakeholders

---

## 📊 Dataset Description

### Synthetic Data: `data/falls_data.csv`

**24 months** of falls data with the following variables:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `month` | Month number | Integer | 1-24 |
| `date` | Month start date | Date | Jan 2023 - Dec 2024 |
| `falls_count` | Number of falls | Integer | 8-26 per month |
| `phase` | Analysis phase | String | Baseline / Post-intervention |

### Data Characteristics

**Realistic patterns included**:
- Baseline mean: ~17.6 falls/month (excluding Month 8)
- Special cause event: Month 8 (flu outbreak, 26 falls)
- Intervention point: Month 13
- Post-intervention mean: ~10.7 falls/month (~40% reduction)
- Natural variation throughout

**Data generation**: Synthetic data generated using R (`set.seed(42)` for reproducibility). The data includes a special cause event and demonstrates a sustained intervention effect. Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

**Purpose**: Load and prepare the falls data for SPC analysis.

---

### Setup

::: {.panel-tabset}

### R

```{r ex2-setup-r}
#| message: false
#| warning: false

library(ggplot2)

# Import falls data
falls_data <- read.csv("data/falls_data.csv", stringsAsFactors = FALSE)

# Convert date column to Date type
falls_data$date <- as.Date(falls_data$date)

# Display first few rows
head(falls_data)
```

### Python

```{python ex2-setup-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt

# Import falls data
falls_data = pd.read_csv("data/falls_data.csv")

# Convert date column to datetime
falls_data['date'] = pd.to_datetime(falls_data['date'])

# Display first few rows
print(falls_data.head())
```

:::

---

### Visualize Time Series

::: {.panel-tabset}

### R

```{r ex2-plot-ts-r}
#| fig-width: 14
#| fig-height: 8

ggplot(falls_data, aes(x = month, y = falls_count)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 12.5, color = "purple", linetype = "dashed", size = 1) +
  annotate("text", x = 12.5, y = max(falls_data$falls_count), 
           label = "Intervention", angle = 90, hjust = -0.1, color = "purple") +
  scale_x_continuous(breaks = seq(0, 24, 3)) +
  labs(
    title = "Monthly Falls Count: Oakwood Care Home",
    subtitle = "24 months (Jan 2023 - Dec 2024)",
    x = "Month",
    y = "Falls Count"
  ) +
  theme_minimal()
```

### Python

```{python ex2-plot-ts-py}

plt.figure(figsize=(14, 8))
plt.plot(falls_data['month'], falls_data['falls_count'], '-o', color='steelblue', linewidth=2, markersize=8)
plt.axvline(12.5, color='purple', linestyle='--', linewidth=2)
plt.text(12.5, falls_data['falls_count'].max(), 'Intervention', rotation=90, ha='right', va='top', color='purple')
plt.xlabel('Month')
plt.ylabel('Falls Count')
plt.title('Monthly Falls Count: Oakwood Care Home\n24 months (Jan 2023 - Dec 2024)')
plt.xticks(range(0, 25, 3))
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Observations**:
- Baseline period (Months 1-12) shows variation around 17-18 falls/month
- Month 8 shows spike (26 falls) - potential special cause
- Post-intervention (Months 13-24) appears lower

---

## Step 2: Baseline SPC

**Purpose**: Establish baseline control limits and assess process stability.

---

### Calculate Control Limits (Including All Data)

::: {.panel-tabset}

### R

```{r ex2-calc-limits-all-r}
# Baseline data (Months 1-12)
baseline_all <- falls_data[falls_data$month <= 12, ]

# Calculate centre line (mean)
centre_line_all <- mean(baseline_all$falls_count)

# Calculate moving range
mr <- abs(diff(baseline_all$falls_count))
mean_mr <- mean(mr)

# Calculate control limits (±3σ)
ucl_all <- centre_line_all + 2.66 * mean_mr
lcl_all <- centre_line_all - 2.66 * mean_mr

cat("Baseline Control Limits (including all data):\n")
cat("  Centre Line:", round(centre_line_all, 2), "\n")
cat("  UCL:", round(ucl_all, 2), "\n")
cat("  LCL:", round(lcl_all, 2), "\n")
```

### Python

```{python ex2-calc-limits-all-py}

# Baseline data
baseline_all = falls_data[falls_data['month'] <= 12]

# Calculate centre line
centre_line_all = baseline_all['falls_count'].mean()

# Calculate moving range
mr = baseline_all['falls_count'].diff().abs().dropna()
mean_mr = mr.mean()

# Calculate control limits
ucl_all = centre_line_all + 2.66 * mean_mr
lcl_all = centre_line_all - 2.66 * mean_mr

print("Baseline Control Limits (including all data):")
print(f"  Centre Line: {centre_line_all:.2f}")
print(f"  UCL: {ucl_all:.2f}")
print(f"  LCL: {lcl_all:.2f}")
```

:::

---

### Identify Special Causes

::: {.panel-tabset}

### R

```{r ex2-identify-special-r}
# Check for points outside limits
baseline_all$outside_limits <- baseline_all$falls_count > ucl_all | baseline_all$falls_count < lcl_all

# Display special causes
special_causes <- baseline_all[baseline_all$outside_limits, c("month", "date", "falls_count")]
print(special_causes)
```

### Python

```{python ex2-identify-special-py}

# Check for points outside limits
baseline_all['outside_limits'] = (baseline_all['falls_count'] > ucl_all) | (baseline_all['falls_count'] < lcl_all)

# Display special causes
special_causes = baseline_all[baseline_all['outside_limits']][['month', 'date', 'falls_count']]
print(special_causes)
```

:::

**Finding**: No points outside control limits when calculated with all baseline data. However, Month 8 (26 falls) shows a notable spike and is a documented special cause (flu outbreak).

**Decision**: Exclude Month 8 from control limit calculation to establish limits based on normal process variation only.

---

### Recalculate Control Limits (Excluding Month 8)

::: {.panel-tabset}

### R

```{r ex2-calc-limits-excl-r}
# Baseline excluding Month 8
baseline_clean <- falls_data[falls_data$month <= 12 & falls_data$month != 8, ]

# Recalculate
centre_line <- mean(baseline_clean$falls_count)
mr_clean <- abs(diff(baseline_clean$falls_count))
mean_mr_clean <- mean(mr_clean)
ucl <- centre_line + 2.66 * mean_mr_clean
lcl <- centre_line - 2.66 * mean_mr_clean

cat("Final Baseline Control Limits (excluding Month 8):\n")
cat("  Centre Line:", round(centre_line, 2), "\n")
cat("  UCL:", round(ucl, 2), "\n")
cat("  LCL:", round(lcl, 2), "\n")

# Baseline parameters calculated
baseline_params <- data.frame(
  centre_line = centre_line,
  ucl = ucl,
  lcl = lcl
)
```

### Python

```{python ex2-calc-limits-excl-py}

# Baseline excluding Month 8
baseline_clean = falls_data[(falls_data['month'] <= 12) & (falls_data['month'] != 8)]

# Recalculate
centre_line = baseline_clean['falls_count'].mean()
mr_clean = baseline_clean['falls_count'].diff().abs().dropna()
mean_mr_clean = mr_clean.mean()
ucl = centre_line + 2.66 * mean_mr_clean
lcl = centre_line - 2.66 * mean_mr_clean

print("Final Baseline Control Limits (excluding Month 8):")
print(f"  Centre Line: {centre_line:.2f}")
print(f"  UCL: {ucl:.2f}")
print(f"  LCL: {lcl:.2f}")

# Baseline parameters calculated
baseline_params = pd.DataFrame({
    'centre_line': [centre_line],
    'ucl': [ucl],
    'lcl': [lcl]
})
```

:::

---

### Plot Baseline Control Chart

::: {.panel-tabset}

### R

```{r ex2-plot-baseline-r}
#| fig-width: 14
#| fig-height: 8

baseline_all$point_type <- ifelse(baseline_all$month == 8, "Special Cause", "Regular")

ggplot(baseline_all, aes(x = month, y = falls_count)) +
  geom_line(color = "gray50", size = 1) +
  geom_point(aes(color = point_type, size = point_type)) +
  scale_color_manual(values = c("Regular" = "steelblue", "Special Cause" = "red")) +
  scale_size_manual(values = c("Regular" = 3, "Special Cause" = 5)) +
  geom_hline(yintercept = centre_line, color = "blue", size = 1) +
  geom_hline(yintercept = ucl, color = "red", linetype = "dashed", size = 1) +
  geom_hline(yintercept = lcl, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = 2, y = centre_line, label = paste("CL =", round(centre_line, 1)), 
           hjust = 0, vjust = -0.5, color = "blue") +
  annotate("text", x = 2, y = ucl, label = paste("UCL =", round(ucl, 1)), 
           hjust = 0, vjust = -0.5, color = "red") +
  annotate("text", x = 2, y = lcl, label = paste("LCL =", round(lcl, 1)), 
           hjust = 0, vjust = 1.5, color = "red") +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  labs(
    title = "Baseline Control Chart (Months 1-12)",
    subtitle = "Control limits calculated excluding Month 8 special cause",
    x = "Month",
    y = "Falls Count",
    color = "Point Type"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Python

```{python ex2-plot-baseline-py}

plt.figure(figsize=(14, 8))

# Plot data
regular = baseline_all[baseline_all['month'] != 8]
special = baseline_all[baseline_all['month'] == 8]

plt.plot(baseline_all['month'], baseline_all['falls_count'], '-', color='gray', linewidth=1, alpha=0.5)
plt.plot(regular['month'], regular['falls_count'], 'o', color='steelblue', markersize=8, label='Regular')
plt.plot(special['month'], special['falls_count'], 'o', color='red', markersize=12, label='Special Cause')

# Control limits
plt.axhline(centre_line, color='blue', linewidth=2, label=f'CL = {centre_line:.1f}')
plt.axhline(ucl, color='red', linestyle='--', linewidth=2, label=f'UCL = {ucl:.1f}')
plt.axhline(lcl, color='red', linestyle='--', linewidth=2, label=f'LCL = {lcl:.1f}')

plt.xlabel('Month')
plt.ylabel('Falls Count')
plt.title('Baseline Control Chart (Months 1-12)\nControl limits calculated excluding Month 8 special cause')
plt.xticks(range(1, 13))
plt.legend()
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Assessment**: Baseline is stable (excluding Month 8 special cause). Ready to extend limits to post-intervention period.

---

## Step 3: Intervention Analysis

**Purpose**: Determine if the falls prevention program reduced falls rates.

---

### Full Control Chart with Intervention

::: {.panel-tabset}

### R

```{r ex2-full-chart-r}
#| fig-width: 14
#| fig-height: 8

falls_data$point_type <- "Regular"
falls_data$point_type[falls_data$month == 8] <- "Special Cause"
falls_data$point_type[falls_data$month >= 13] <- "Post-Intervention"

ggplot(falls_data, aes(x = month, y = falls_count)) +
  geom_line(color = "gray50", size = 1) +
  geom_point(aes(color = point_type, size = point_type)) +
  scale_color_manual(values = c("Regular" = "steelblue", 
                                 "Special Cause" = "red",
                                 "Post-Intervention" = "darkgreen")) +
  scale_size_manual(values = c("Regular" = 3, "Special Cause" = 5, "Post-Intervention" = 3)) +
  geom_hline(yintercept = centre_line, color = "blue", size = 1) +
  geom_hline(yintercept = ucl, color = "red", linetype = "dashed", size = 1) +
  geom_hline(yintercept = lcl, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = 12.5, color = "purple", linetype = "solid", size = 1.5) +
  annotate("text", x = 12.5, y = max(falls_data$falls_count), 
           label = "Intervention Start", angle = 90, hjust = -0.1, 
           color = "purple", fontface = "bold") +
  scale_x_continuous(breaks = seq(0, 24, 3)) +
  labs(
    title = "Full Control Chart: Baseline and Post-Intervention",
    subtitle = "Baseline limits extended to post-intervention period",
    x = "Month",
    y = "Falls Count",
    color = "Point Type"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Python

```{python ex2-full-chart-py}

plt.figure(figsize=(14, 8))

baseline = falls_data[falls_data['month'] <= 12]
post = falls_data[falls_data['month'] >= 13]
special = falls_data[falls_data['month'] == 8]

plt.plot(falls_data['month'], falls_data['falls_count'], '-', color='gray', linewidth=1, alpha=0.5)
plt.plot(baseline[baseline['month'] != 8]['month'], baseline[baseline['month'] != 8]['falls_count'], 
         'o', color='steelblue', markersize=8, label='Baseline')
plt.plot(special['month'], special['falls_count'], 'o', color='red', markersize=12, label='Special Cause')
plt.plot(post['month'], post['falls_count'], 'o', color='darkgreen', markersize=8, label='Post-Intervention')

plt.axhline(centre_line, color='blue', linewidth=2)
plt.axhline(ucl, color='red', linestyle='--', linewidth=2)
plt.axhline(lcl, color='red', linestyle='--', linewidth=2)
plt.axvline(12.5, color='purple', linestyle='-', linewidth=2)

plt.xlabel('Month')
plt.ylabel('Falls Count')
plt.title('Full Control Chart: Baseline and Post-Intervention\nBaseline limits extended to post-intervention period')
plt.xticks(range(0, 25, 3))
plt.legend()
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

---

### Detect Points Outside Limits

::: {.panel-tabset}

### R

```{r ex2-detect-outside-r}
post_intervention <- falls_data[falls_data$month >= 13, ]

outside_lcl <- post_intervention[post_intervention$falls_count < lcl, ]

cat("Post-Intervention Points Below LCL (indicating improvement):\n")
print(outside_lcl[, c("month", "date", "falls_count")])
cat("\nTotal:", nrow(outside_lcl), "of", nrow(post_intervention), "months\n")
```

### Python

```{python ex2-detect-outside-py}

post_intervention = falls_data[falls_data['month'] >= 13]

outside_lcl = post_intervention[post_intervention['falls_count'] < lcl]

print("Post-Intervention Points Below LCL (indicating improvement):")
print(outside_lcl[['month', 'date', 'falls_count']])
print(f"\nTotal: {len(outside_lcl)} of {len(post_intervention)} months")
```

:::

---

### Apply Detection Rule 4

**Rule 4**: 8+ consecutive points one side of centre line indicates process shift.

::: {.panel-tabset}

### R

```{r ex2-rule4-r}
post_intervention$below_centre <- post_intervention$falls_count < centre_line

cat("Rule 4: 8+ consecutive points one side of centre line\n\n")
cat("Post-intervention points below baseline centre line:\n")
cat("  Total:", sum(post_intervention$below_centre), "of", nrow(post_intervention), "\n")

consecutive_below <- rle(post_intervention$below_centre)
max_run <- max(consecutive_below$lengths[consecutive_below$values])

cat("  Longest consecutive run below centre:", max_run, "\n")

if (max_run >= 8) {
  cat("\n✅ Rule 4 TRIGGERED: Process has shifted\n")
  cat("   Strong evidence of sustained improvement\n")
}
```

### Python

```{python ex2-rule4-py}

post_intervention['below_centre'] = post_intervention['falls_count'] < centre_line

print("Rule 4: 8+ consecutive points one side of centre line\n")
print(f"Post-intervention points below baseline centre line: {post_intervention['below_centre'].sum()} of {len(post_intervention)}")

from itertools import groupby
runs = [(k, len(list(g))) for k, g in groupby(post_intervention['below_centre'])]
max_run = max([length for value, length in runs if value], default=0)

print(f"Longest consecutive run below centre: {max_run}")

if max_run >= 8:
    print("\n✅ Rule 4 TRIGGERED: Process has shifted")
    print("   Strong evidence of sustained improvement")
```

:::

---

### Statistical Comparison

::: {.panel-tabset}

### R

```{r ex2-statistical-comparison-r}
baseline_data <- falls_data[falls_data$phase == "Baseline" & falls_data$month != 8, ]
post_data <- falls_data[falls_data$phase == "Post-intervention", ]

cat("Statistical Comparison:\n\n")
cat("Baseline (excluding Month 8):\n")
cat("  Mean:", round(mean(baseline_data$falls_count), 2), "falls/month\n")
cat("  SD:", round(sd(baseline_data$falls_count), 2), "\n")

cat("\nPost-Intervention:\n")
cat("  Mean:", round(mean(post_data$falls_count), 2), "falls/month\n")
cat("  SD:", round(sd(post_data$falls_count), 2), "\n")

reduction <- mean(baseline_data$falls_count) - mean(post_data$falls_count)
pct_reduction <- 100 * reduction / mean(baseline_data$falls_count)

cat("\nDifference:\n")
cat("  Absolute reduction:", round(reduction, 2), "falls/month\n")
cat("  Percentage reduction:", round(pct_reduction, 1), "%\n")

t_result <- t.test(baseline_data$falls_count, post_data$falls_count)
cat("\nTwo-sample t-test: p =", format.pval(t_result$p.value, digits = 3), "\n")
```

### Python

```{python ex2-statistical-comparison-py}

from scipy import stats

baseline_data = falls_data[(falls_data['phase'] == 'Baseline') & (falls_data['month'] != 8)]
post_data = falls_data[falls_data['phase'] == 'Post-intervention']

print("Statistical Comparison:\n")
print(f"Baseline: Mean = {baseline_data['falls_count'].mean():.2f}, SD = {baseline_data['falls_count'].std():.2f}")
print(f"Post-Intervention: Mean = {post_data['falls_count'].mean():.2f}, SD = {post_data['falls_count'].std():.2f}")

reduction = baseline_data['falls_count'].mean() - post_data['falls_count'].mean()
pct_reduction = 100 * reduction / baseline_data['falls_count'].mean()

print(f"\nReduction: {reduction:.2f} falls/month ({pct_reduction:.1f}%)")

t_stat, p_value = stats.ttest_ind(baseline_data['falls_count'], post_data['falls_count'])
print(f"Two-sample t-test: p = {p_value:.4f}")
```

:::

---

## Step 4: Summary and Recommendations

### Intervention Effect Detected

**Evidence**:
- ✅ Multiple points below baseline LCL
- ✅ Rule 4 triggered (12 consecutive below centre)
- ✅ Statistical significance (p < 0.001)
- ✅ Clinically significant reduction (~7 falls/month, 40%)
- ✅ Improvement sustained throughout 12-month post-intervention period

**Evidence Strength**: **STRONG**

---

### Recommendations

1. **Continue program**: Falls prevention program has demonstrably reduced falls
2. **Ongoing monitoring**: Continue monthly SPC charting
3. **Share best practice**: Document program components for sector-wide rollout
4. **Investigate further improvements**: Can we reduce further?

---

### Limitations

- Single care home (generalizability unknown)
- No control group (external factors possible)
- 12-month follow-up (longer-term sustainability unknown)
- Month 8 special cause excluded from limits

---

**SPC Analysis Complete** ✅

---

### Summary

This example demonstrates a complete SPC workflow for intervention evaluation, including:

✅ Time-series visualization and data preparation  
✅ Baseline control chart with special cause handling  
✅ Control limit calculation (I-chart method)  
✅ Intervention effect detection using SPC rules  
✅ Statistical confirmation and sustainability assessment  
✅ Stakeholder-ready reporting and recommendations  

**For full QA documentation, see**: `docs/qa-checklist.md`, `docs/decisions-log.md`
