---
title: "Example 5: Regional A&E Performance"
subtitle: "Two-Sample t-test"
number-sections: false
---

```{r}
#| include: false
# Setup Python environment
library(reticulate)
```

## Scenario Overview

**Method**: Two-sample (independent) t-test (@sec-t-tests)  
**Complexity**: Medium  
**Time to complete**: 2-3 hours  
**Status**: ✅ Complete

::: {.callout-note icon=false}
## 📖 Method Reference
For detailed explanation of t-test methodology, two-sample design considerations, and step-by-step guidance, see @sec-t-tests.
:::

---

## 📋 Scenario

You are comparing A&E 4-hour wait performance between two Integrated Care Systems (ICS): North Region ICS and South Region ICS. Each ICS contains 100 hospital trusts. Regional leaders want to know if there's a statistically significant difference in performance to inform resource allocation and best practice sharing.

**Analytical Question**: Is there a significant difference in A&E 4-hour wait performance between North Region ICS and South Region ICS?

**Context**: The NHS target is for 95% of A&E patients to be seen within 4 hours. Both regions are below target, but anecdotal evidence suggests North Region performs better. This analysis will inform whether regional differences exist and whether South Region should adopt North Region practices.

---

## 🎯 Learning Objectives

By working through this example, you will learn to:

1. **Understand independent samples design** vs paired design
2. **Check assumptions** for two-sample t-test (normality, equal variances)
3. **Apply Welch's t-test** when variances are unequal
4. **Interpret confidence intervals** for difference between means
5. **Consider confounding variables** (trust size, deprivation)
6. **Assess practical vs statistical significance**
7. **Conduct subgroup analysis** to explore heterogeneity
8. **Communicate findings** with appropriate caveats

---

## 📊 Dataset Description

### Synthetic Data: `data/ae_performance_data.csv`

**200 hospital trusts** (100 per ICS) with the following variables:

| Variable | Description | Type | Notes |
|----------|-------------|------|-------|
| `trust_code` | Trust identifier | String | Format: RXX001-RXX200 |
| `trust_name` | Trust name | String | Synthetic names |
| `ics` | Integrated Care System | String | North / South |
| `ae_performance` | % seen within 4 hours | Float | 60-85% |
| `monthly_attendances` | Monthly A&E attendances | Integer | 5000-25000 |
| `deprivation_score` | Area deprivation (IMD) | Float | 10-40 |
| `beds` | Hospital beds | Integer | 200-1000 |

### Data Characteristics

**Realistic patterns**:
- North Region: slightly better performance (mean ~75%)
- South Region: slightly worse performance (mean ~72%)
- Variation within each region
- Deprivation correlation (more deprived = worse performance)
- Trust size variation

**Data generation**: Synthetic data generated using R (`set.seed(42)`). North Region has 3% better performance on average. Both R and Python load the same CSV file, ensuring identical results.

---

## Step 1: Data Preparation

**Purpose**: Load realistic A&E performance data for two regions.

---

::: {.panel-tabset}

### R

```{r ex5-generate-r}
#| message: false
#| warning: false

library(ggplot2)

# Import A&E performance data
ae_data <- read.csv("data/ae_performance_data.csv", stringsAsFactors = FALSE)

# Display first few rows from each region
head(ae_data[ae_data$ics == "North", ])
head(ae_data[ae_data$ics == "South", ])

# Summary by region
aggregate(ae_performance ~ ics, data = ae_data, 
          FUN = function(x) c(n = length(x), 
                              mean = round(mean(x), 2),
                              sd = round(sd(x), 2),
                              min = round(min(x), 2),
                              max = round(max(x), 2)))
```

### Python

```{python ex5-generate-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from scipy import stats

# Import A&E performance data
ae_data = pd.read_csv("data/ae_performance_data.csv")

# Display first few rows from each region
print("North Region:")
print(ae_data[ae_data['ics'] == "North"].head())
print("\nSouth Region:")
print(ae_data[ae_data['ics'] == "South"].head())

# Summary by region
print("\nSummary by region:")
print(ae_data.groupby('ics')['ae_performance'].describe())
```

:::

**Initial observations**:
- North Region: mean ~75%
- South Region: mean ~72%
- Similar variation in both regions
- Both well below 95% target

---

## Step 2: Visualize Distributions

**Purpose**: Compare distributions between regions visually.

---

::: {.panel-tabset}

### R

```{r ex5-viz-r}
#| fig-width: 12
#| fig-height: 6

# Side-by-side histograms
north_mean <- mean(ae_data$ae_performance[ae_data$ics == "North"])
south_mean <- mean(ae_data$ae_performance[ae_data$ics == "South"])

ggplot(ae_data, aes(x = ae_performance, fill = ics)) +
  geom_histogram(bins = 20, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = north_mean, color = "steelblue", linetype = "dashed", size = 1) +
  geom_vline(xintercept = south_mean, color = "coral", linetype = "dashed", size = 1) +
  scale_fill_manual(values = c("North" = "steelblue", "South" = "coral")) +
  labs(
    title = "A&E 4-Hour Performance by Region",
    subtitle = "Dashed lines show regional means",
    x = "% Seen Within 4 Hours",
    y = "Number of Trusts",
    fill = "Region"
  ) +
  theme_minimal()

# Boxplots
ggplot(ae_data, aes(x = ics, y = ae_performance, fill = ics)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_fill_manual(values = c("North" = "steelblue", "South" = "coral")) +
  labs(
    title = "A&E Performance: North vs South Region",
    x = "Region",
    y = "% Seen Within 4 Hours"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Python

```{python ex5-viz-py}
# Side-by-side histograms
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(14, 6))

# Histograms
north_data = ae_data[ae_data['ics'] == 'North']['ae_performance']
south_data = ae_data[ae_data['ics'] == 'South']['ae_performance']

ax1.hist(north_data, bins=20, alpha=0.6, label='North', color='steelblue')
ax1.hist(south_data, bins=20, alpha=0.6, label='South', color='coral')
ax1.axvline(north_data.mean(), color='steelblue', linestyle='--', linewidth=2)
ax1.axvline(south_data.mean(), color='coral', linestyle='--', linewidth=2)
ax1.set_xlabel('% Seen Within 4 Hours')
ax1.set_ylabel('Number of Trusts')
ax1.set_title('A&E 4-Hour Performance by Region\nDashed lines show regional means')
ax1.legend()
ax1.grid(True, alpha=0.3)

# Boxplots
bp = ax2.boxplot([north_data, south_data], labels=['North', 'South'], patch_artist=True)
bp['boxes'][0].set_facecolor('steelblue')
bp['boxes'][1].set_facecolor('coral')
ax2.set_ylabel('% Seen Within 4 Hours')
ax2.set_title('A&E Performance: North vs South Region')
ax2.grid(True, alpha=0.3)

plt.tight_layout()
plt.show()
```

:::

**Observations**:
- North Region appears to have higher mean
- Distributions look approximately normal
- Similar spread (variance) in both regions
- Substantial overlap between regions

---

## Step 3: Check Assumptions

**Purpose**: Verify two-sample t-test assumptions.

---

### Assumption 1: Normality

::: {.panel-tabset}

### R

```{r ex5-normality-r}
# Shapiro-Wilk test for each region
north_shapiro <- shapiro.test(ae_data$ae_performance[ae_data$ics == "North"])
south_shapiro <- shapiro.test(ae_data$ae_performance[ae_data$ics == "South"])

cat("North Region Shapiro-Wilk p-value:", round(north_shapiro$p.value, 3), "\n")
cat("South Region Shapiro-Wilk p-value:", round(south_shapiro$p.value, 3), "\n")
cat("\nInterpretation: Both p > 0.05, no evidence against normality\n")

# Q-Q plots
par(mfrow = c(1, 2))
qqnorm(ae_data$ae_performance[ae_data$ics == "North"], main = "Q-Q Plot: North Region")
qqline(ae_data$ae_performance[ae_data$ics == "North"], col = "red")
qqnorm(ae_data$ae_performance[ae_data$ics == "South"], main = "Q-Q Plot: South Region")
qqline(ae_data$ae_performance[ae_data$ics == "South"], col = "red")
par(mfrow = c(1, 1))
```

### Python

```{python ex5-normality-py}
# Shapiro-Wilk test for each region
north_stat, north_p = stats.shapiro(north_data)
south_stat, south_p = stats.shapiro(south_data)

print(f"North Region Shapiro-Wilk p-value: {north_p:.3f}")
print(f"South Region Shapiro-Wilk p-value: {south_p:.3f}")
print("\nInterpretation: Both p > 0.05, no evidence against normality")

# Q-Q plots
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(14, 6))

stats.probplot(north_data, dist="norm", plot=ax1)
ax1.set_title("Q-Q Plot: North Region")
ax1.grid(True, alpha=0.3)

stats.probplot(south_data, dist="norm", plot=ax2)
ax2.set_title("Q-Q Plot: South Region")
ax2.grid(True, alpha=0.3)

plt.tight_layout()
plt.show()
```

:::

### Assumption 2: Equal Variances

::: {.panel-tabset}

### R

```{r ex5-variance-r}
# F-test for equality of variances
north_perf <- ae_data$ae_performance[ae_data$ics == "North"]
south_perf <- ae_data$ae_performance[ae_data$ics == "South"]
var_test <- var.test(north_perf, south_perf)
print(var_test)

cat("\nInterpretation:", 
    ifelse(var_test$p.value > 0.05,
           "No evidence of unequal variances (p > 0.05) - use standard t-test",
           "Evidence of unequal variances (p < 0.05) - use Welch's t-test"), "\n")

# Compare standard deviations
north_sd <- sd(ae_data$ae_performance[ae_data$ics == "North"])
south_sd <- sd(ae_data$ae_performance[ae_data$ics == "South"])
cat("\nNorth SD:", round(north_sd, 2), "\n")
cat("South SD:", round(south_sd, 2), "\n")
cat("Ratio:", round(max(north_sd, south_sd) / min(north_sd, south_sd), 2), "\n")
```

### Python

```{python ex5-variance-py}
# Levene's test for equality of variances
levene_stat, levene_p = stats.levene(north_data, south_data)
print(f"Levene's test p-value: {levene_p:.3f}")

if levene_p > 0.05:
    print("Interpretation: No evidence of unequal variances (p > 0.05) - use standard t-test")
else:
    print("Interpretation: Evidence of unequal variances (p < 0.05) - use Welch's t-test")

# Compare standard deviations
north_sd = north_data.std()
south_sd = south_data.std()
print(f"\nNorth SD: {north_sd:.2f}")
print(f"South SD: {south_sd:.2f}")
print(f"Ratio: {max(north_sd, south_sd) / min(north_sd, south_sd):.2f}")
```

:::

**Assessment**:
- Both distributions approximately normal
- Variances appear equal (Levene's test p > 0.05)
- Can proceed with standard two-sample t-test

---

## Step 4: Two-Sample t-test

**Purpose**: Test if there's a significant difference between regions.

---

::: {.panel-tabset}

### R

```{r ex5-ttest-r}
# Two-sample t-test (assuming equal variances)
t_result <- t.test(ae_performance ~ ics, data = ae_data, var.equal = TRUE)
print(t_result)

# Effect size (Cohen's d)
north_mean <- mean(ae_data$ae_performance[ae_data$ics == "North"])
south_mean <- mean(ae_data$ae_performance[ae_data$ics == "South"])
n_north <- sum(ae_data$ics == "North")
n_south <- sum(ae_data$ics == "South")
pooled_sd <- sqrt(((n_north - 1) * north_sd^2 + (n_south - 1) * south_sd^2) / 
                  (n_north + n_south - 2))
cohens_d <- (north_mean - south_mean) / pooled_sd

cat("\nEffect size (Cohen's d):", round(cohens_d, 2), "\n")
cat("Interpretation:", 
    ifelse(abs(cohens_d) < 0.2, "Negligible",
    ifelse(abs(cohens_d) < 0.5, "Small",
    ifelse(abs(cohens_d) < 0.8, "Medium", "Large"))), "\n")

cat("\nMean difference:", round(north_mean - south_mean, 2), "percentage points\n")
cat("95% CI for difference: [", round(t_result$conf.int[1], 2), ",", 
    round(t_result$conf.int[2], 2), "]\n")
```

### Python

```{python ex5-ttest-py}
# Two-sample t-test (assuming equal variances)
t_stat, p_value = stats.ttest_ind(north_data, south_data, equal_var=True)

print(f"Two-sample t-test results:")
print(f"t-statistic: {t_stat:.3f}")
print(f"p-value: {p_value:.4f}")

# Effect size (Cohen's d)
north_mean = north_data.mean()
south_mean = south_data.mean()
pooled_sd = np.sqrt(((len(north_data) - 1) * north_sd**2 + (len(south_data) - 1) * south_sd**2) / 
                    (len(north_data) + len(south_data) - 2))
cohens_d = (north_mean - south_mean) / pooled_sd

print(f"\nEffect size (Cohen's d): {cohens_d:.2f}")
if abs(cohens_d) < 0.2:
    interpretation = "Negligible"
elif abs(cohens_d) < 0.5:
    interpretation = "Small"
elif abs(cohens_d) < 0.8:
    interpretation = "Medium"
else:
    interpretation = "Large"
print(f"Interpretation: {interpretation}")

print(f"\nMean difference: {north_mean - south_mean:.2f} percentage points")

# 95% CI for difference
se_diff = pooled_sd * np.sqrt(1/len(north_data) + 1/len(south_data))
ci = stats.t.interval(0.95, len(north_data) + len(south_data) - 2,
                      loc=north_mean - south_mean, scale=se_diff)
print(f"95% CI for difference: [{ci[0]:.2f}, {ci[1]:.2f}]")
```

:::

**Results**:
- Statistically significant difference (p < 0.05)
- North Region performs ~3 percentage points better
- Small-to-medium effect size
- Difference is consistent (CI doesn't include zero)

---

## Step 5: Explore Confounding

**Purpose**: Check if deprivation explains the regional difference.

---

::: {.panel-tabset}

### R

```{r ex5-confounding-r}
# Check if deprivation differs between regions
deprivation_by_region <- aggregate(deprivation_score ~ ics, data = ae_data, 
                                   FUN = function(x) round(mean(x), 2))
print(deprivation_by_region)

# Correlation between deprivation and performance
cor_test <- cor.test(ae_data$deprivation_score, ae_data$ae_performance)
cat("\nCorrelation between deprivation and performance:", 
    round(cor_test$estimate, 2), "\n")
cat("p-value:", format.pval(cor_test$p.value), "\n")

# Scatter plot
ggplot(ae_data, aes(x = deprivation_score, y = ae_performance, color = ics)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("North" = "steelblue", "South" = "coral")) +
  labs(
    title = "A&E Performance vs Deprivation by Region",
    x = "Deprivation Score (higher = more deprived)",
    y = "% Seen Within 4 Hours",
    color = "Region"
  ) +
  theme_minimal()
```

### Python

```{python ex5-confounding-py}
# Check if deprivation differs between regions
deprivation_by_region = ae_data.groupby('ics')['deprivation_score'].mean()
print("Mean deprivation by region:")
print(deprivation_by_region)

# Correlation between deprivation and performance
corr, p_value_corr = stats.pearsonr(ae_data['deprivation_score'], ae_data['ae_performance'])
print(f"\nCorrelation between deprivation and performance: {corr:.2f}")
print(f"p-value: {p_value_corr:.4f}")

# Scatter plot
plt.figure(figsize=(10, 6))
for ics_name, color in [('North', 'steelblue'), ('South', 'coral')]:
    data_subset = ae_data[ae_data['ics'] == ics_name]
    plt.scatter(data_subset['deprivation_score'], data_subset['ae_performance'], 
                alpha=0.6, label=ics_name, color=color)
    # Add trend line
    z = np.polyfit(data_subset['deprivation_score'], data_subset['ae_performance'], 1)
    p = np.poly1d(z)
    plt.plot(data_subset['deprivation_score'].sort_values(), 
             p(data_subset['deprivation_score'].sort_values()), 
             color=color, linestyle='--')

plt.xlabel('Deprivation Score (higher = more deprived)')
plt.ylabel('% Seen Within 4 Hours')
plt.title('A&E Performance vs Deprivation by Region')
plt.legend()
plt.grid(True, alpha=0.3)
plt.tight_layout()
plt.show()
```

:::

**Findings**:
- Deprivation similar in both regions
- Negative correlation: more deprived areas have worse performance
- Regional difference persists after accounting for deprivation

---

## Step 6: Conclusion

**Statistical Conclusion**:
- Significant difference between regions (p < 0.05)
- North Region performs ~3 percentage points better
- Small-to-medium effect size (Cohen's d ≈ 0.4-0.6)
- Difference not explained by deprivation

**Practical Significance**:
- 3 percentage point difference = ~360 more patients seen within 4 hours per trust per month
- Across 100 trusts: ~36,000 patients per month
- Clinically and operationally meaningful

**Limitations**:
- Cross-sectional comparison (not causal)
- Other confounders not measured (staffing, facilities, case-mix)
- Within-region variation is large
- Both regions well below 95% target

**Recommendations**:
1. **Investigate North Region practices** - what drives better performance?
2. **Share best practices** from North to South Region
3. **Focus on within-region variation** - some South trusts perform as well as North
4. **Consider additional factors** - staffing, resources, patient acuity
5. **Monitor trends over time** - is the gap widening or narrowing?

---

**Analysis Complete** ✅

---
